---
title: "NHANES Oral Microbiome GOLD Database Integration & Genus Mapping"
author: "Terry Cho"
date: "2025-01-19"
output: html_document
---

```{r knitr_init, echo=FALSE, warning=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

# Global options
options(max.print = 75)
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  eval = TRUE,
  cache = FALSE,  # Disable cache for reproducibility
  echo = TRUE,
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(width = 85)
```

# NHANES Oral Microbiome GOLD Database Integration Pipeline

> **This pipeline integrates GOLD database microbial phenotype data with NHANES oral microbiome genus-level data**

## ⚠️ **Prerequisites**

**Before running this script, ensure you have completed:**

1. **Chapter 0:** Data transformation (creates transformed microbiome tables)
2. **Chapter 1:** Association analysis (optional, but recommended for context)
3. **Chapter 2:** Database preprocessing and phyloseq object creation

## **Setup & Libraries**

```{r libraries, message=FALSE}
# Set global chunk options for RMarkdown
knitr::opts_chunk$set(echo = TRUE)

# List of required packages
required_packages <- c(
  "data.table", "dplyr", "dbplyr", "RSQLite", "readr", "tidyr", 
  "tibble", "purrr", "phyloseq", "kableExtra"
)

# Ensure CRAN mirror is set globally
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install missing packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, dependencies = TRUE)
}

# Load all required packages
invisible(lapply(required_packages, library, character.only = TRUE))

cat("✅ All required packages loaded successfully\n")
```

## **Setup Paths and Database Connection**

```{r setup_paths}
base_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho"

# Output paths - corrected directory structure
viz_out_path <- file.path(base_path, "results/analyses_results/03_gold_db_microbial_phenotype_out/figures_out")
intermediate_files_path <- file.path(base_path, "results/analyses_results/03_gold_db_microbial_phenotype_out/intermediate")

# Input paths
phyloseq_obj_files_path <- file.path(base_path, "results/analyses_results/02_preprocess_db_n_phyloseq_out/intermediate")
db_path <- file.path(base_path, "data/00_nhanes_omp_transformed_db/nhanes_oral_transformed_complete_processed.sqlite")
gold_db_subset_path <- file.path(base_path, "data/00_GOLDdb/goldData_ubiome_selected.csv")

# Create directories if they don't exist
dir.create(viz_out_path, recursive = TRUE, showWarnings = FALSE)
dir.create(intermediate_files_path, recursive = TRUE, showWarnings = FALSE)

# Verify required files exist
if (!file.exists(db_path)) {
  stop("❌ Processed database not found. Please run Chapter 2, Step 1 first:\n",
       "Rscript scripts/2_preprocess_db_n_phyloseq/process_demographics_complete.R")
}

if (!file.exists(gold_db_subset_path)) {
  stop("❌ GOLD database file not found at: ", gold_db_subset_path)
}

if (!file.exists(file.path(phyloseq_obj_files_path, "ubiome_relative.rds"))) {
  stop("❌ phyloseq objects not found. Please run Chapter 2, Step 3 first:\n",
       "scripts/2_preprocess_db_n_phyloseq/ubiome_phyloseq_object_creation_cho.Rmd")
}

# Connect to database
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)

cat("✅ Connected to processed database:", db_path, "\n")
cat("✅ GOLD database file found:", gold_db_subset_path, "\n")
cat("✅ phyloseq objects directory:", phyloseq_obj_files_path, "\n")
cat("✅ Output directory:", intermediate_files_path, "\n")
```

## **Database Verification**

```{r verify_database}
# ---- Global Save PDF Function ----
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {
  pdf(
    file = file.path(viz_out_path, file_name),
    width = width_mm / 25.4, height = height_mm / 25.4,
    family = "Helvetica", colormodel = "rgb",
    pointsize = 5, useDingbats = FALSE
  )
}

# List all tables in the database
all_tables <- dbListTables(con)

# Identify tables for oral microbiome cycles (F and G)
years_F_or_G <- all_tables[grepl("_F$|_G$", all_tables)]

cat("Total tables in database:", length(all_tables), "\n",
    "Tables with F/G cycles:", length(years_F_or_G), "\n")

# Check for specific data tables
table_names <- tbl(con, "table_names_epcf")
specific_tables <- table_names %>%
  filter(`Data.File.Name` %like% '%d_outcome_mcq_f%' | 
         `Data.File.Name` %like% '%d_outcome_mcq_g%' | 
         `Data.File.Name` %like% '%d_outcome_oh_f%' | 
         `Data.File.Name` %like% '%d_outcome_oh_g%') %>%
  collect()

# Check variable metadata
varnames <- tbl(con, "variable_names_epcf")
total_variables <- varnames %>% summarise(count = n()) %>% collect()

cat("Specific outcome tables found:", nrow(specific_tables), "\n",
    "Total variables in metadata:", total_variables$count, "\n",
    "✅ Database verification complete\n")
```

# **Import and Process GOLD Database**

## **Step 1: Load and Clean GOLD Database**

```{r load_gold_db}
cat("Loading and processing GOLD database...\n")

# Load CSV as character to avoid factor conversion issues
gold_csv <- fread(gold_db_subset_path,
  colClasses = "character",
  encoding = "UTF-8"
)

cat("Original GOLD database size:", nrow(gold_csv), "rows ×", ncol(gold_csv), "columns\n")

# Clean text encoding and standardize missing values
gold_csv[] <- lapply(gold_csv, function(x) {
  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x[tolower(x) %in% c("na", "unclassified", "undefined", "unknown", "")] <- NA_character_
  x
})

# Fix column names
names(gold_csv) <- iconv(names(gold_csv), from = "latin1", to = "UTF-8")
names(gold_csv) <- gsub("<a1>", "°", names(gold_csv))
names(gold_csv) <- gsub("^ORGANISM ", "", names(gold_csv))
names(gold_csv) <- gsub(" ", "_", names(gold_csv))

cat("Column names after cleaning:\n")
print(names(gold_csv))

# Select only the desired columns for analysis
required_columns <- c("NCBI_KINGDOM", "GOLD_PHYLUM", "GENUS", "SPECIES", "BIOTIC_RELATIONSHIPS",
                     "OXYGEN_REQUIREMENT", "METABOLISM", "ENERGY_SOURCES", "GRAM_STAIN",
                     "SAMPLE_COLLECTION_DATE", "SALINITY", "CELL_SHAPE", "MOTILITY", "SPORULATION",
                     "TEMPERATURE_RANGE")

# Check which columns exist
existing_columns <- intersect(required_columns, names(gold_csv))
missing_columns <- setdiff(required_columns, names(gold_csv))

if (length(missing_columns) > 0) {
  cat("⚠️ Missing columns:", paste(missing_columns, collapse = ", "), "\n")
}

cat("Selecting", length(existing_columns), "available columns for analysis\n")
gold_csv <- gold_csv[, ..existing_columns]

cat("✅ GOLD database loaded and cleaned:", nrow(gold_csv), "rows ×", ncol(gold_csv), "columns\n")
```

## **Step 2: Create Categorical Scores and Clean Data**

```{r clean_gold_data}
cat("Creating categorical scores and cleaning data...\n")

# Clean up and group values for OXYGEN_REQUIREMENT
gold_csv[, OXYGEN_REQUIREMENT := fcase(
  OXYGEN_REQUIREMENT %in% c("Anaerobe", "Obligate anaerobe"), "Anaerobe",
  OXYGEN_REQUIREMENT %in% c("Aerobe", "Obligate aerobe"), "Aerobe",
  OXYGEN_REQUIREMENT %in% c("Facultative", "Facultative anaerobe", "Facultative aerobe"), "Facultative",
  OXYGEN_REQUIREMENT %in% c("Microaerophilic"), "Microaerophilic",
  default = NA_character_
)]

# Create OXYGEN_SCORE
gold_csv[, OXYGEN_SCORE := fcase(
  OXYGEN_REQUIREMENT == "Anaerobe", 0,
  OXYGEN_REQUIREMENT == "Facultative", 0.5,
  OXYGEN_REQUIREMENT == "Microaerophilic", 0.25,
  OXYGEN_REQUIREMENT == "Aerobe", 1,
  default = NA_real_
)]

# Clean up GRAM_STAIN
gold_csv[, GRAM_STAIN := fcase(
  GRAM_STAIN == "Gram-", "Gram-",
  GRAM_STAIN == "Gram+", "Gram+",
  default = NA_character_
)]

# Create STAIN_SCORE
gold_csv[, STAIN_SCORE := fcase(
  GRAM_STAIN == "Gram-", 0,
  GRAM_STAIN == "Gram+", 1,
  default = NA_real_
)]

# Clean up MOTILITY
gold_csv[, MOTILITY := fcase(
  MOTILITY == "Nonmotile", "Nonmotile",
  MOTILITY %in% c("Motile", "Chemotactic"), "Motile",
  default = NA_character_
)]

# Create MOTILITY_SCORE
gold_csv[, MOTILITY_SCORE := fcase(
  MOTILITY == "Nonmotile", 0,
  MOTILITY == "Motile", 1,
  default = NA_real_
)]

# Clean up SPORULATION
gold_csv[, SPORULATION := fcase(
  SPORULATION == "Nonsporulating", "Nonsporulating",
  SPORULATION == "Sporulating", "Sporulating",
  default = NA_character_
)]

# Create SPORULATION_SCORE
gold_csv[, SPORULATION_SCORE := fcase(
  SPORULATION == "Nonsporulating", 0,
  SPORULATION == "Sporulating", 1,
  default = NA_real_
)]

cat("✅ Categorical scores created successfully\n")

# Display summary of scores
score_summary <- gold_csv[, .(
  OXYGEN_mean = round(mean(OXYGEN_SCORE, na.rm = TRUE), 3),
  OXYGEN_na = sum(is.na(OXYGEN_SCORE)),
  STAIN_mean = round(mean(STAIN_SCORE, na.rm = TRUE), 3),
  STAIN_na = sum(is.na(STAIN_SCORE)),
  MOTILITY_mean = round(mean(MOTILITY_SCORE, na.rm = TRUE), 3),
  MOTILITY_na = sum(is.na(MOTILITY_SCORE)),
  SPORULATION_mean = round(mean(SPORULATION_SCORE, na.rm = TRUE), 3),
  SPORULATION_na = sum(is.na(SPORULATION_SCORE))
)]

cat("Score summary:\n")
print(score_summary)
```

## **Step 3: Define Aggregation Functions**

```{r aggregation_functions}
cat("Defining aggregation functions...\n")

# Define column categories
skip_cols <- c("OXYGEN_SCORE", "STAIN_SCORE", "MOTILITY_SCORE", "SPORULATION_SCORE")
text_cols <- c("METABOLISM", "ENERGY_SOURCES")
char_cols <- names(gold_csv)[sapply(gold_csv, is.character)]
categorical_cols <- setdiff(char_cols, c(text_cols, skip_cols))
features <- setdiff(names(gold_csv), c(skip_cols, "GENUS"))

# Assign types to each feature
feature_types <- sapply(features, function(f) {
  if (f %in% text_cols) "text"
  else if (f %in% categorical_cols) "categorical"
  else "skip"
})

cat("Feature types:\n")
print(feature_types)

# Summarization functions
summarize_categorical <- function(values, threshold = 0.7) {
  total_count <- length(values)
  valid_values <- na.omit(values)
  Non_NA_count <- as.integer(length(valid_values))
  
  if (Non_NA_count == 0) {
    return(list(
      aggregation_value = "unknown (0%)", 
      agreement = "none", 
      Non_NA_count = 0L, 
      NA_count = total_count
    ))
  }
  
  freq <- table(valid_values) / Non_NA_count
  aggregation_value <- paste(names(freq), " (", round(freq * 100, 1), "%)", sep = "", collapse = ", ")
  
  if (length(unique(valid_values)) == 1) {
    agreement <- "high"
  } else if (max(freq) >= threshold) {
    agreement <- "medium"
  } else {
    agreement <- "low"
  }
  
  NA_count <- total_count - Non_NA_count
  list(
    aggregation_value = aggregation_value, 
    agreement = agreement, 
    Non_NA_count = Non_NA_count, 
    NA_count = NA_count
  )
}

summarize_text <- function(values, n_top = 3) {
  total_count <- length(values)
  valid_values <- na.omit(values)
  Non_NA_count <- as.integer(length(valid_values))
  
  if (Non_NA_count == 0) {
    return(list(
      aggregation_value = "unknown (0%)", 
      Non_NA_count = 0L, 
      NA_count = total_count
    ))
  }
  
  tokens <- unlist(strsplit(valid_values, "[|,;]"))
  tokens <- trimws(tokens)
  freq <- table(tokens) / length(tokens)
  top_terms <- head(sort(freq, decreasing = TRUE), n_top)
  aggregation_value <- paste(names(top_terms), " (", round(top_terms * 100, 1), "%)", sep = "", collapse = ", ")
  
  NA_count <- total_count - Non_NA_count
  list(
    aggregation_value = aggregation_value, 
    Non_NA_count = Non_NA_count, 
    NA_count = NA_count
  )
}

summarize_index <- function(scores) {
  total_count <- length(scores)
  scores <- na.omit(scores)
  Non_NA_count <- as.integer(length(scores))
  
  if (Non_NA_count == 0) {
    return(list(
      aggregation_value = "unknown (0%)", 
      Non_NA_count = 0L, 
      NA_count = total_count
    ))
  }
  
  index <- mean(scores)  # Store raw mean value
  NA_count <- total_count - Non_NA_count
  list(
    aggregation_value = index, 
    Non_NA_count = Non_NA_count, 
    NA_count = NA_count
  )
}

summarize_feature <- function(data, feature, type) {
  if (type == "skip") return(NULL)
  
  values <- data[[feature]]
  
  if (type == "categorical") {
    summary <- summarize_categorical(values)
    return(data.table(
      feature = feature, 
      aggregation_value = summary$aggregation_value, 
      agreement = summary$agreement, 
      Non_NA_count = summary$Non_NA_count, 
      NA_count = summary$NA_count
    ))
  } else if (type == "text") {
    summary <- summarize_text(values)
    return(data.table(
      feature = feature, 
      aggregation_value = summary$aggregation_value, 
      agreement = "medium", 
      Non_NA_count = summary$Non_NA_count, 
      NA_count = summary$NA_count
    ))
  } else {
    return(NULL)
  }
}

cat("✅ Aggregation functions defined successfully\n")
```

## **Step 4: Aggregate Features by Genus**

```{r aggregate_by_genus}
cat("Aggregating features by genus...\n")

# Check if GENUS column exists
if (!"GENUS" %in% names(gold_csv)) {
  stop("❌ GENUS column not found in GOLD database")
}

# Count unique genera before aggregation
unique_genera_count <- gold_csv[, uniqueN(GENUS, na.rm = TRUE)]
cat("Unique genera in GOLD database:", unique_genera_count, "\n")

# Aggregate all features by genus and calculate indices
gold_db_genus <- gold_csv[, {
  feature_summaries <- lapply(features, function(f) {
    type <- feature_types[f]
    summarize_feature(.SD, f, type)
  })
  
  # Calculate indices
  indices <- list(
    data.table(
      feature = "STAIN_INDEX", 
      aggregation_value = summarize_index(.SD$STAIN_SCORE)$aggregation_value, 
      agreement = "", 
      Non_NA_count = summarize_index(.SD$STAIN_SCORE)$Non_NA_count, 
      NA_count = summarize_index(.SD$STAIN_SCORE)$NA_count
    ),
    data.table(
      feature = "OXYGEN_INDEX", 
      aggregation_value = summarize_index(.SD$OXYGEN_SCORE)$aggregation_value, 
      agreement = "", 
      Non_NA_count = summarize_index(.SD$OXYGEN_SCORE)$Non_NA_count, 
      NA_count = summarize_index(.SD$OXYGEN_SCORE)$NA_count
    ),
    data.table(
      feature = "MOTILITY_INDEX", 
      aggregation_value = summarize_index(.SD$MOTILITY_SCORE)$aggregation_value, 
      agreement = "", 
      Non_NA_count = summarize_index(.SD$MOTILITY_SCORE)$Non_NA_count, 
      NA_count = summarize_index(.SD$MOTILITY_SCORE)$NA_count
    ),
    data.table(
      feature = "SPORULATION_INDEX", 
      aggregation_value = summarize_index(.SD$SPORULATION_SCORE)$aggregation_value, 
      agreement = "", 
      Non_NA_count = summarize_index(.SD$SPORULATION_SCORE)$Non_NA_count, 
      NA_count = summarize_index(.SD$SPORULATION_SCORE)$NA_count
    )
  )
  
  # Combine feature summaries and indices
  rbind(
    do.call(rbind, feature_summaries),
    do.call(rbind, indices),
    fill = TRUE
  )
}, by = "GENUS"]

# Display summary
cat("✅ GOLD database aggregated by genus:\n",
    "  Genera:", gold_db_genus[, uniqueN(GENUS)], "\n",
    "  Features per genus:", gold_db_genus[, .N, by = GENUS][, mean(N)], "\n")

# Save the aggregated GOLD database
gold_db_genus_file <- file.path(intermediate_files_path, "gold_db_genus.csv")
fwrite(gold_db_genus, gold_db_genus_file)

cat("✅ Saved aggregated GOLD database to:", gold_db_genus_file, "\n",
    "\nSample of aggregated GOLD database:\n")
print(head(gold_db_genus, 10))
```

# **Load phyloseq Objects and Create Genus Mapping**

## **Step 5: Load phyloseq Object**

```{r load_phyloseq}
cat("Loading phyloseq objects...\n")

# Load ubiome relative abundance phyloseq object
ubiome_relative_file <- file.path(phyloseq_obj_files_path, "ubiome_relative.rds")

if (!file.exists(ubiome_relative_file)) {
  stop("❌ phyloseq object not found at:", ubiome_relative_file)
}

ubiome_relative <- readRDS(ubiome_relative_file)

cat("✅ phyloseq object loaded successfully:\n",
    "  Samples:", nsamples(ubiome_relative), "\n",
    "  Taxa:", ntaxa(ubiome_relative), "\n",
    "  Sample variables:", length(sample_variables(ubiome_relative)), "\n",
    "  Taxonomic ranks:", paste(rank_names(ubiome_relative), collapse = ", "), "\n")

# Verify tax_table exists
if (is.null(tax_table(ubiome_relative))) {
  stop("❌ Taxonomy table not found in phyloseq object")
}

cat("✅ Taxonomy table verified\n")
```

## **Step 6: Create Genus Mapping from phyloseq**

```{r create_genus_mapping}
cat("Creating genus mapping from phyloseq taxonomy...\n")

# Extract taxonomy table
tax_data <- tax_table(ubiome_relative)

# Check if Genus column exists
if (!"Genus" %in% colnames(tax_data)) {
  stop("❌ Genus column not found in taxonomy table")
}

# Create genus mapping dataframe
ubiome_genus_mapping <- data.frame(
  otu = rownames(tax_data),
  Genus = tax_data[, "Genus"],
  stringsAsFactors = FALSE
)

cat("Original genus mapping created:", nrow(ubiome_genus_mapping), "OTUs\n",
    "Unique genera:", length(unique(ubiome_genus_mapping$Genus)), "\n",
    "\nSample of original genus mapping:\n")
print(head(ubiome_genus_mapping, 10))
```

## **Step 7: Parse Genus Names for Matching**

```{r parse_genus_names}
cat("Parsing genus names for better matching...\n")

# Function to parse core genus from various nomenclature patterns
parse_genus <- function(genus_name) {
  if (is.na(genus_name)) return("unclassified")
  
  # Start with the original name
  genus <- genus_name
  
  # Special case: "Incertae Sedis" -> "unclassified"
  if (genus_name == "Incertae Sedis") {
    return("unclassified")
  }
  
  # Special case: "Escherichia/Shigella" -> "unclassified"
  if (genus_name == "Escherichia/Shigella") {
    return("unclassified")
  }
  
  # 1. "[<Genus>] <specific_detail> group" or "[<Genus>] <specific_detail>" -> <Genus>
  if (grepl("^\\[.*\\]", genus_name)) {
    genus <- sub("^\\[([^]]+)\\].*$", "\\1", genus_name)
  }
  
  # 2. "Candidatus <Genus>" -> <Genus>
  if (grepl("^Candidatus\\s+", genus)) {
    genus <- sub("^Candidatus\\s+(\\S+).*$", "\\1", genus)
  }
  
  # 3. "<Genus> sensu stricto <number>" -> <Genus>
  if (grepl("sensu stricto", genus)) {
    genus <- sub("^(.*?)\\s+sensu stricto\\s+\\d+.*$", "\\1", genus)
  }
  
  # 4. Handle group patterns
  if (grepl("\\s+", genus) && !grepl("sensu stricto", genus)) {
    # Patterns with "group" suffix
    if (grepl("\\s+.*group$", genus)) {
      # Handle numeric prefixes like "AEGEAN-169", "CPla-4", "dgA-11", "GKS98"
      if (grepl("^[A-Z]+\\d+[-]?\\d*", genus)) {
        genus <- sub("^(.*?\\d+[-]?\\d*)\\s+.*$", "\\1", genus)
      } else {
        genus <- sub("^(.*?)\\s+.*group$", "\\1", genus)
      }
    } else {
      # Handle "<Genus> <number>" or "<Genus> <specific_detail>"
      genus <- sub("^(.*?)\\s+(\\S+)$", "\\1", genus)
    }
  }
  
  # 5. Special case for "aceae" families with "group" suffix
  if (grepl("aceae\\s+.*group$", genus)) {
    genus <- sub("^(.*aceae).*$", "\\1", genus)
  }
  
  # Return the parsed genus, fallback to original if problematic
  if (genus == "" || (genus == genus_name && !grepl("^[A-Za-z0-9-]+$", genus))) {
    genus <- genus_name
  }
  
  return(genus)
}

# Apply parsing function
ubiome_genus_mapping$parsed_genus <- sapply(ubiome_genus_mapping$Genus, parse_genus)

# Summary of parsing results
# Handle NA values in comparison properly
genus_comparison <- !is.na(ubiome_genus_mapping$Genus) & 
                   !is.na(ubiome_genus_mapping$parsed_genus) & 
                   ubiome_genus_mapping$Genus != ubiome_genus_mapping$parsed_genus

parsing_summary <- data.frame(
  original_unique = length(unique(ubiome_genus_mapping$Genus[!is.na(ubiome_genus_mapping$Genus)])),
  parsed_unique = length(unique(ubiome_genus_mapping$parsed_genus[!is.na(ubiome_genus_mapping$parsed_genus)])),
  unclassified_count = sum(ubiome_genus_mapping$parsed_genus == "unclassified", na.rm = TRUE),
  changed_count = sum(genus_comparison, na.rm = TRUE)
)

cat("✅ Genus parsing completed:\n",
    "  Original unique genera:", parsing_summary$original_unique, "\n",
    "  Parsed unique genera:", parsing_summary$parsed_unique, "\n",
    "  Unclassified genera:", parsing_summary$unclassified_count, "\n",
    "  Names changed by parsing:", parsing_summary$changed_count, "\n",
    "\nExamples of genus name parsing:\n")

# Filter for changed names, handling NAs properly
if (sum(genus_comparison, na.rm = TRUE) > 0) {
  changed_examples <- ubiome_genus_mapping[genus_comparison, c("Genus", "parsed_genus")]
  examples <- unique(changed_examples)
  examples <- head(examples, 10)
  print(examples)
} else {
  cat("No genus names were changed during parsing\n")
}
```

# **Merge NHANES and GOLD Database**

## **Step 8: Merge phyloseq Genera with GOLD Database**

```{r merge_databases}
cat("Merging phyloseq genera with GOLD database...\n")

# Define features to merge from GOLD database
features_to_merge <- c(
  "BIOTIC_RELATIONSHIPS", "OXYGEN_REQUIREMENT", "METABOLISM", "ENERGY_SOURCES", 
                       "GRAM_STAIN", "SAMPLE_COLLECTION_DATE", "SALINITY", "CELL_SHAPE", "MOTILITY", 
                       "SPORULATION", "TEMPERATURE_RANGE", "STAIN_INDEX", "OXYGEN_INDEX", 
  "MOTILITY_INDEX", "SPORULATION_INDEX"
)

# Pivot GOLD database to wide format for merging
cat("Pivoting GOLD database to wide format...\n")
gold_db_genus_wide <- dcast(
  gold_db_genus, 
  GENUS ~ feature, 
  value.var = "aggregation_value", 
  fun.aggregate = function(x) if (length(x) > 0) x[1] else NA_character_
)

# Check which features are available
available_features <- intersect(features_to_merge, names(gold_db_genus_wide))
missing_features <- setdiff(features_to_merge, names(gold_db_genus_wide))

if (length(missing_features) > 0) {
  cat("⚠️ Missing features in GOLD database:", paste(missing_features, collapse = ", "), "\n")
}

cat("Available features for merging:", length(available_features), "\n")

# Select available columns for merging
gold_columns <- c("GENUS", available_features)
gold_db_genus_subset <- gold_db_genus_wide[, ..gold_columns]

cat("GOLD database subset for merging:", nrow(gold_db_genus_subset), "genera ×", ncol(gold_db_genus_subset), "features\n")

# Merge with ubiome_genus_mapping using parsed_genus
ubiome_genus_mapping_complete <- merge(
  ubiome_genus_mapping,
  gold_db_genus_subset,
  by.x = "parsed_genus",
  by.y = "GENUS",
  all.x = TRUE,  # Keep all rows from ubiome_genus_mapping
  all.y = FALSE
)

# Convert to data.table for efficient handling
setDT(ubiome_genus_mapping_complete)

cat("✅ Merge completed:", nrow(ubiome_genus_mapping_complete), "rows\n")

# Ensure all expected features exist (add NA columns if missing)
missing_cols <- setdiff(available_features, names(ubiome_genus_mapping_complete))
if (length(missing_cols) > 0) {
  ubiome_genus_mapping_complete[, (missing_cols) := NA_character_]
  cat("Added missing columns:", paste(missing_cols, collapse = ", "), "\n")
}

# Replace "unknown (0%)" with NA for cleaner data
ubiome_genus_mapping_complete <- ubiome_genus_mapping_complete[, 
  lapply(.SD, function(x) fifelse(x == "unknown (0%)", NA_character_, x))
]

cat("✅ Data cleaning completed\n")
```

## **Step 9: Analyze Mapping Results**

```{r analyze_mapping}
cat("Analyzing mapping results...\n")

# Basic statistics
total_genera <- nrow(ubiome_genus_mapping_complete)
unique_original_genera <- length(unique(ubiome_genus_mapping_complete$Genus))
unique_parsed_genera <- length(unique(ubiome_genus_mapping_complete$parsed_genus))

# Count annotations
if (length(available_features) > 0) {
  # Check for any non-NA annotations
  annotated_rows <- ubiome_genus_mapping_complete[, 
    any(!is.na(.SD)), .SDcols = available_features, by = .(otu, Genus)
  ]
  annotated_count <- sum(annotated_rows$V1)
  
  # Count unclassified
  unclassified_count <- sum(ubiome_genus_mapping_complete$parsed_genus == "unclassified")
  
  # Count completely unannotated (all features NA)
  na_counts <- ubiome_genus_mapping_complete[, 
    lapply(.SD, is.na), .SDcols = available_features
  ]
  unannotated_count <- sum(rowSums(na_counts) == length(available_features))
  
} else {
  annotated_count <- 0
  unclassified_count <- sum(ubiome_genus_mapping_complete$parsed_genus == "unclassified")
  unannotated_count <- total_genera
}

# Calculate coverage statistics
total_excluding_unclassified <- total_genera - unclassified_count
annotation_coverage <- ifelse(total_genera > 0, round((annotated_count / total_genera) * 100, 2), 0)
annotation_coverage_excl_unclassified <- ifelse(
  total_excluding_unclassified > 0, 
  round((annotated_count / total_excluding_unclassified) * 100, 2), 
  0
)

# Report results
cat("\nMapping Results Summary:\n",
    "  Total OTUs:", total_genera, "\n",
    "  Unique original genera:", unique_original_genera, "\n",
    "  Unique parsed genera:", unique_parsed_genera, "\n",
    "  OTUs with GOLD annotations:", annotated_count, "\n",
    "  OTUs parsed as 'unclassified':", unclassified_count, "\n",
    "  OTUs with no annotations:", unannotated_count, "\n",
    "  Annotation coverage (%):", annotation_coverage, "%\n",
    "  Annotation coverage excluding unclassified (%):", annotation_coverage_excl_unclassified, "%\n")

# Identify unmapped genera (excluding unclassified)
if (length(available_features) > 0) {
  unmapped_genera <- ubiome_genus_mapping_complete[
    rowSums(na_counts) == length(available_features) & 
    parsed_genus != "unclassified", 
    unique(Genus)
  ]
  
  if (length(unmapped_genera) > 0) {
    cat("\n⚠️ Genera without GOLD annotations (first 20):\n")
    cat(paste(head(unmapped_genera, 20), collapse = ", "), "\n")
    if (length(unmapped_genera) > 20) {
      cat("... and", length(unmapped_genera) - 20, "more\n")
    }
} else {
    cat("\n✅ All classifiable genera have GOLD annotations\n")
}
}
```

## **Step 10: Save Results**

```{r save_results}
cat("💾 Saving mapping results...\n")

# Save the complete mapping
output_file <- file.path(intermediate_files_path, "ubiome_genus_mapping_complete.csv")
fwrite(ubiome_genus_mapping_complete, output_file)

# Create summary statistics file
summary_stats <- data.table(
  metric = c(
    "total_otus",
    "unique_original_genera", 
    "unique_parsed_genera",
    "annotated_otus",
    "unclassified_otus",
    "unannotated_otus",
    "annotation_coverage_percent",
    "annotation_coverage_excl_unclassified_percent"
  ),
  value = c(
    total_genera,
    unique_original_genera,
    unique_parsed_genera,
    annotated_count,
    unclassified_count,
    unannotated_count,
    annotation_coverage,
    annotation_coverage_excl_unclassified
  )
)

summary_file <- file.path(intermediate_files_path, "mapping_summary_stats.csv")
fwrite(summary_stats, summary_file)

cat("✅ Complete genus mapping saved to:", output_file, "\n",
    "✅ Summary statistics saved to:", summary_file, "\n",
    "\nFinal mapping structure:\n")
str(ubiome_genus_mapping_complete)

# Close database connection
dbDisconnect(con)
cat("\n🔌 Database connection closed\n")
```

# ✅ **Completion Summary**

```{r completion_summary}
cat("✅ NHANES-GOLD Database Integration Complete!\n\n",
    "Results Created:\n",
    "  1. Aggregated GOLD database by genus:", file.path(intermediate_files_path, "gold_db_genus.csv"), "\n",
    "  2. Complete genus mapping:", file.path(intermediate_files_path, "ubiome_genus_mapping_complete.csv"), "\n",
    "  3. Summary statistics:", file.path(intermediate_files_path, "mapping_summary_stats.csv"), "\n\n",
    "Key Statistics:\n",
    "  - Total OTUs processed:", total_genera, "\n",
    "  - OTUs with GOLD annotations:", annotated_count, "(", annotation_coverage, "%)\n",
    "  - Available GOLD features:", length(available_features), "\n",
    "  - Unique genera in GOLD database:", gold_db_genus[, uniqueN(GENUS)], "\n\n",
    "🔗 Integration with Analysis Pipeline:\n",
    "  - Use ubiome_genus_mapping_complete for microbial phenotype analysis\n",
    "  - Combine with phyloseq objects for functional annotation\n",
    "  - Apply to association results for biological interpretation\n\n",
    "📖 Next Steps:\n",
    "  1. Proceed to Chapter 4: Downstream analysis with GOLD annotations\n",
    "  2. Perform functional enrichment analysis\n",
    "  3. Interpret association results using microbial phenotypes\n")
```

## 📚 **Variable Names for Downstream Compatibility**

### **Key Variables Maintained for Pipeline Compatibility:**

- **`ubiome_genus_mapping_complete`** - Main output table with GOLD annotations
- **`gold_db_genus`** - Aggregated GOLD database by genus
- **`features_to_merge`** - List of GOLD features integrated
- **File naming** - Maintained original file names for downstream scripts

### **Available GOLD Features:**
```{r display_features}
if (exists("available_features")) {
  cat("Available GOLD features for analysis:\n")
  for (i in seq_along(available_features)) {
    cat("  ", i, ". ", available_features[i], "\n", sep = "")
  }
} else {
  cat("No GOLD features available\n")
}
```

**This pipeline successfully integrates NHANES oral microbiome data with GOLD database microbial phenotypes for comprehensive functional analysis!**