---
title: "ubiome_res_collect"
author: "Terry Cho"
date: "2025-01-25"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)
manager_packages <- c("BiocManager", "remotes") # Install managers if needed
invisible(lapply(manager_packages, function(pkg) {if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)}))

# CRAN packages (load in this order)
cran_packages <- c(
  # Core data infrastructure
  "Matrix", "data.table", "purrr",
  "readr", "tidyr", "stringr", "forcats", "rlang",
  "DBI",
  # Legacy / meta
  "tidyverse","tibble", "dplyr",
  # Reporting
  "knitr", "kableExtra",
  "matrixTests",
  # Visualization
  "ggplot2", "viridis", "egg", "ggrepel", "ggthemes", "patchwork",
  "GGally","dtw","ggbeeswarm",
  "gridExtra", "pheatmap", "RColorBrewer", "plotly", "scales",
  "circlize", "grid", "vegan", "extrafont"
)
# Bioconductor packages
bioconductor_packages <- c(
  "phyloseq",       # microbiome data structures & plotting
  "ComplexHeatmap", # advanced heatmaps
  "grafify",        # genomics plotting API
  "BiocParallel",   # parallel backends
  "sva"            # batch‐correction workflows
)

github_packages <- c("chiragjp/nhanespewas") 

install_if_missing <- function(pkg, type = "cran") {
  name <- sub(".*/", "", pkg)
  if (!requireNamespace(name, quietly = TRUE)) {
    switch(type, cran = install.packages(pkg),
                bioc = BiocManager::install(pkg, ask = FALSE),
                github = remotes::install_github(pkg))}}
invisible(lapply(cran_packages, install_if_missing, type = "cran"))
invisible(lapply(bioconductor_packages, install_if_missing, type = "bioc"))
invisible(lapply(github_packages, install_if_missing, type = "github"))

# Load all libraries
all_packages <- c(cran_packages, bioconductor_packages, sub(".*/", "", github_packages))
invisible(lapply(all_packages, library, character.only = TRUE))

cat("\nLoaded package versions:\n") # Display loaded versions
for (pkg in all_packages) {version <- tryCatch(as.character(packageVersion(pkg)), error = function(e) "N/A")
                           cat(sprintf("  %s: %s\n", pkg, version))}
# Environment setup
register(MulticoreParam(4))
```

```{R}
# ---- Paths ----
viz_out_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/results/analyses_results/5_age_analyses_out"

# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",  # Use Helvetica font  
    colormodel = "rgb",    # Ensure color consistency  
    pointsize = 5,         # Default base point size  
    useDingbats = FALSE    # Prevent font embedding issues  
  )  
}

# ---- Paths ----
intermediate_files_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/results/analyses_results/4_association_phyloseq_analyses_out/intermediate"
# ---- Database Connection ----
db_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/data/00_nhanes_omp_abundance_db/nhanes_031725.sqlite"
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)

table_description <- tbl(con, "table_names_epcf")
variable_description <- tbl(con, "variable_names_epcf")
variable_description
```


Below is the full R code to analyze the oral microbiome data across different age groups using hierarchical clustering and parallel coordinate plots, as specified in your query. The code includes filtering genus-level OTUs with a prevalence of at least 0.1 (non-zero in at least 10% of samples), reducing the data size by computing mean relative abundances grouped into 5-year incremental age groups from 14-19 up to 80-85, and performing the subsequent clustering and visualization steps.

```{r}
# Set seed for reproducibility
set.seed(42)

# Load necessary libraries
library(phyloseq)
library(dplyr)
# library(GGally)
# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",     # Use Helvetica font  
    colormodel = "rgb",       # Ensure color consistency  
    pointsize = 5,            # Default base point size  
    useDingbats = FALSE       # Prevent font embedding issues  
  )  
}  

# Load the ubiome relative abundance phyloseq object
ubiome_relative <- readRDS(file.path(intermediate_files_path, "ubiome_relative_none_updated.rds"))
cat("Phyloseq object loaded from:", file.path(intermediate_files_path, "ubiome_relative_none_updated.rds"), "\n")

# Step 1: Aggregate OTUs to genus level
ubiome_genus <- tax_glom(ubiome_relative, taxrank = "Genus")
cat("Aggregated to genus level. Number of genera:", ntaxa(ubiome_genus), "\n")

# Step 2: Filter genera with prevalence >= 0.1 (non-zero in at least 10% of samples)
prevalence_threshold <- 0.1 * nsamples(ubiome_genus)
ubiome_filtered <- filter_taxa(ubiome_genus, function(x) sum(x > 0) >= prevalence_threshold, TRUE)
cat("Filtered genera with prevalence >= 0.1. Remaining genera:", ntaxa(ubiome_filtered), "\n")

# Step 3: Define age groups (14-19, 20-24, ..., 80-85) and add to sample_data
age_breaks <- c(14, seq(20, 85, by = 5))
age_labels <- c("14-19", paste(seq(20, 75, by = 5), seq(24, 79, by = 5), sep = "-"), "80-85")
sample_data(ubiome_filtered)$age_group <- cut(
  sample_data(ubiome_filtered)$Age,
  breaks = age_breaks,
  labels = age_labels,
  right = FALSE
)
cat("Age groups defined and added to sample data.\n")

```

```{R}
# Set seed for reproducibility
set.seed(42)

# Step 4: Compute mean relative abundance per age group
otu_mat <- as(otu_table(ubiome_filtered), "matrix")  # Taxa as rows, samples as columns
sample_df <- as(sample_data(ubiome_filtered), "data.frame")
otu_t <- t(otu_mat)  # Transpose so samples are rows, taxa are columns
otu_df <- as.data.frame(otu_t)
otu_df$age_group <- sample_df$age_group
mean_abundance <- otu_df %>%
  group_by(age_group) %>%
  summarise(across(everything(), mean, na.rm = TRUE))
mean_abundance_mat <- as.matrix(mean_abundance[, -1])  # Remove age_group column
rownames(mean_abundance_mat) <- mean_abundance$age_group
cat("Computed mean relative abundance per age group. Dimensions:", dim(mean_abundance_mat), "\n")

# Step 5: Identify differentially abundant genera across age groups using Kruskal-Wallis test
p_values <- apply(otu_mat, 1, function(x) kruskal.test(x ~ sample_df$age_group)$p.value)
adj_p_values <- p.adjust(p_values, method = "BH")  # Benjamini-Hochberg correction
significant_genera <- names(which(adj_p_values < 0.05))
cat("Identified", length(significant_genera), "significant genera (adjusted p < 0.05).\n")

# Step 6: Subset mean abundance matrix to significant genera
mean_abundance_sig <- mean_abundance_mat[, significant_genera, drop = FALSE]

# Step 7: Standardize the mean abundance profiles (mean = 0, SD = 1) for each genus
mean_abundance_sig_t <- t(mean_abundance_sig)  # Genera as rows, age groups as columns
mean_abundance_scaled <- t(scale(t(mean_abundance_sig_t)))  # Scale each genus across age groups
cat("Standardized mean abundance profiles.\n")

# Step 8: Perform hierarchical clustering using Ward's linkage
dist_mat <- dist(mean_abundance_scaled, method = "euclidean")
hc <- hclust(dist_mat, method = "ward.D2")
cat("Performed hierarchical clustering with Ward's method.\n")

# Step 9: Determine optimal number of clusters using adjusted silhouette method
# Use a more sensitive approach for microbiome data
sil_scores <- numeric()
sil_scores_penalized <- numeric()

for (k in 2:min(20, nrow(mean_abundance_scaled))) {  # Extended range, but not more than data points
  clusters_temp <- cutree(hc, k = k)
  sil <- silhouette(clusters_temp, dist_mat)
  sil_scores[k - 1] <- mean(sil)  # Use mean function directly on silhouette object
  
  # Apply penalty for very small clusters (less than 3% of data)
  min_cluster_size <- nrow(mean_abundance_scaled) * 0.03
  cluster_sizes <- table(clusters_temp)
  small_cluster_penalty <- sum(cluster_sizes < min_cluster_size) * 0.1
  
  # Penalize if silhouette width is very low
  sil_penalty <- ifelse(mean(sil) < 0.2, 0.2, 0)
  
  # Combined penalty
  sil_scores_penalized[k - 1] <- mean(sil) - small_cluster_penalty - sil_penalty
}

# Find the optimal k using the penalized silhouette scores
optimal_k <- which.max(sil_scores_penalized) + 1

# Apply biological constraints
optimal_k <- max(optimal_k, 4)  # Minimum 4 clusters for microbiome data
optimal_k <- min(optimal_k, 15)  # Maximum 15 clusters to avoid overfitting

cat("Silhouette analysis results:\n")
cat("  Raw silhouette scores (k=2-20):", round(sil_scores, 3), "\n")
cat("  Penalized silhouette scores:", round(sil_scores_penalized, 3), "\n")
cat("  Optimal k (penalized):", optimal_k, "clusters\n")
cat("  Final optimal k (constrained):", optimal_k, "clusters\n")

# Cut the tree into the optimal number of clusters
clusters_numeric <- cutree(hc, k = optimal_k)
cat("Using", optimal_k, "clusters determined by adjusted silhouette analysis.\n")

# Step 10: Convert numeric cluster assignments to character cluster names
# Create dynamic cluster mapping based on optimal number of clusters
cluster_letters <- letters[1:min(optimal_k, 26)]  # Use letters a-z (max 26 clusters)
cluster_mapping <- setNames(cluster_letters, as.character(1:optimal_k))

# Convert numeric clusters to character clusters
clusters <- factor(cluster_mapping[as.character(clusters_numeric)], 
                  levels = cluster_letters)

# Define cluster colors (dynamic based on number of clusters)
# Use a predefined color palette that can handle up to 26 clusters
cluster_color_palette <- c(
  "#FF80DA", "#FF9289", "#F3AC00", "#89CC00", "#00DCA9", "#00D8EF", 
  "#9CB1FF", "#D69EFF", "#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4",
  "#FFEAA7", "#DDA0DD", "#98D8C8", "#F7DC6F", "#BB8FCE", "#85C1E9",
  "#F8C471", "#82E0AA", "#F1948A", "#85C1E9", "#D7BDE2", "#A9DFBF",
  "#F9E79F", "#AED6F1"
)
cluster_colors <- setNames(cluster_color_palette[1:optimal_k], cluster_letters)

# Step 11: Prepare data for parallel coordinate plot
mean_abundance_df <- as.data.frame(mean_abundance_scaled)
mean_abundance_df$genus <- rownames(mean_abundance_df)
mean_abundance_df$cluster <- clusters

# Extract phylum information for each genus
tax_table_df <- as.data.frame(tax_table(ubiome_filtered))
mean_abundance_df$phylum <- tax_table_df[mean_abundance_df$genus, "Phylum"]

# Define phylum colors (STRICT DEFINITION - NEVER DEVIATE)
phylum_colors <- c(
  "Firmicutes"           = "#E69F00",  # orange
  "Bacteroidetes"        = "#56B4E9",  # blue
  "Actinobacteria"       = "#009E73",  # green
  "Proteobacteria"       = "#F0E442",  # yellow
  "Fusobacteria"         = "#CC6677",  # dark blue
  "Spirochaetae"         = "#D55E00",  # reddish-orange
  "Cyanobacteria"        = "#CC79A7",  # purple
  "Acidobacteria"        = "#999999",  # grey
  "Candidate division SR1" = "#AD7700",# brown
  "Planctomycetes"       = "#332288",  # deep indigo
  "Saccharibacteria"     = "#44AA99",  # teal
  "Synergistetes"        = "#88CCEE",  # light blue
  "Tenericutes"          = "#117733",  # dark green
  "unclassified"         = "#DDDDDD",  # very light grey
  "NA"                   = "#DDDDDD",  # very light grey
  "Verrucomicrobia"      = "#0072B2"   # reddish pink
)

# Step 12: Create parallel coordinate plot colored by cluster
p_cluster <- ggparcoord(
  data = mean_abundance_df,
  columns = 1:(ncol(mean_abundance_df) - 3),  # Exclude genus, cluster, phylum columns
  groupColumn = "cluster",
  scale = "globalminmax",  # Use standardized values as is
  title = "Parallel Coordinate Plot of Genus Abundance Trends by Age Group (Cluster)",
  alphaLines = 0.5  # Set transparency
) +
  scale_color_manual(values = cluster_colors) +  # Apply custom cluster colors
  egg::theme_article() +  # White background, minimal grid
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  ) +
  labs(x = "Age Group", y = "Standardized Relative Abundance", color = "Cluster")
print(p_cluster)
cat("Generated parallel coordinate plot colored by cluster.\n")

# Step 13: Create parallel coordinate plot colored by phylum
p_phylum <- ggparcoord(
  data = mean_abundance_df,
  columns = 1:(ncol(mean_abundance_df) - 3),  # Exclude genus, cluster, phylum columns
  groupColumn = "phylum",
  scale = "globalminmax",  # Use standardized values as is
  title = "Parallel Coordinate Plot of Genus Abundance Trends by Age Group (Phylum)",
  alphaLines = 0.5  # Set transparency
) +
  scale_color_manual(values = phylum_colors) +  # Apply custom phylum colors
  egg::theme_article() +  # White background, minimal grid
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  ) +
  labs(x = "Age Group", y = "Standardized Relative Abundance", color = "Phylum")
print(p_phylum)
cat("Generated parallel coordinate plot colored by phylum.\n")

# Step 14: Save both plots as PDFs
# Save plot colored by cluster
save_pdf_figure("15.1_parallel_coord_plot_cluster.pdf", width_mm = 55, height_mm = 45)
print(p_cluster)
dev.off()
cat("Saved parallel coordinate plot colored by cluster as PDF.\n")

# Save plot colored by phylum
save_pdf_figure("15.2_parallel_coord_plot_phylum.pdf", width_mm = 75, height_mm = 45)
print(p_phylum)
dev.off()
cat("Saved parallel coordinate plot colored by phylum as PDF.\n")
```


```{R}
# Set seed for reproducibility
set.seed(42)

# Required libraries
library(phyloseq)      # For microbiome data handling
library(dplyr)         # For data manipulation
library(GGally)        # For parallel coordinate plots
library(cluster)       # For silhouette analysis
library(Rtsne)         # For t-SNE
library(pheatmap)      # For heatmaps
library(ggplot2)       # For plotting
library(cowplot)       # For arranging multiple plots
library(egg)           # For theme_article()

# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",  # Use Helvetica font  
    colormodel = "rgb",    # Ensure color consistency  
    pointsize = 3,         # Default base point size  
    useDingbats = FALSE    # Prevent font embedding issues  
  )  
}  

# Step 4: Compute mean relative abundance per age group
otu_mat <- as(otu_table(ubiome_filtered), "matrix")  # Taxa as rows, samples as columns
sample_df <- as(sample_data(ubiome_filtered), "data.frame")
otu_t <- t(otu_mat)  # Transpose so samples are rows, taxa are columns
otu_df <- as.data.frame(otu_t)
otu_df$age_group <- sample_df$age_group
mean_abundance <- otu_df %>%
  group_by(age_group) %>%
  summarise(across(everything(), mean, na.rm = TRUE))
mean_abundance_mat <- as.matrix(mean_abundance[, -1])  # Remove age_group column
rownames(mean_abundance_mat) <- mean_abundance$age_group
cat("Computed mean relative abundance per age group. Dimensions:", dim(mean_abundance_mat), "\n")

# Step 5: Identify differentially abundant genera across age groups using Kruskal-Wallis test
p_values <- apply(otu_mat, 1, function(x) kruskal.test(x ~ sample_df$age_group)$p.value)
adj_p_values <- p.adjust(p_values, method = "BH")  # Benjamini-Hochberg correction
significant_genera <- names(which(adj_p_values < 0.05))
cat("Identified", length(significant_genera), "significant genera (adjusted p < 0.05).\n")

# Step 6: Subset mean abundance matrix to significant genera
mean_abundance_sig <- mean_abundance_mat[, significant_genera, drop = FALSE]

# Step 7: Standardize the mean abundance profiles (mean = 0, SD = 1) for each genus
mean_abundance_sig_t <- t(mean_abundance_sig)  # Genera as rows, age groups as columns
mean_abundance_scaled <- t(scale(t(mean_abundance_sig_t)))  # Scale each genus across age groups
cat("Standardized mean abundance profiles.\n")

# Step 8: Perform hierarchical clustering using Ward's linkage
dist_mat <- dist(mean_abundance_scaled, method = "euclidean")
hc <- hclust(dist_mat, method = "ward.D2")
cat("Performed hierarchical clustering with Ward's method.\n")

# Step 9: Determine optimal number of clusters using adjusted silhouette method
# Use a more sensitive approach for microbiome data
sil_scores <- numeric()
sil_scores_penalized <- numeric()

for (k in 2:min(20, nrow(mean_abundance_scaled))) {  # Extended range, but not more than data points
  clusters_temp <- cutree(hc, k = k)
  sil <- silhouette(clusters_temp, dist_mat)
  sil_scores[k - 1] <- mean(sil)  # Use mean function directly on silhouette object
  
  # Apply penalty for very small clusters (less than 3% of data)
  min_cluster_size <- nrow(mean_abundance_scaled) * 0.03
  cluster_sizes <- table(clusters_temp)
  small_cluster_penalty <- sum(cluster_sizes < min_cluster_size) * 0.1
  
  # Penalize if silhouette width is very low
  sil_penalty <- ifelse(mean(sil) < 0.2, 0.2, 0)
  
  # Combined penalty
  sil_scores_penalized[k - 1] <- mean(sil) - small_cluster_penalty - sil_penalty
}

# Find the optimal k using the penalized silhouette scores
optimal_k <- which.max(sil_scores_penalized) + 1

# Apply biological constraints
optimal_k <- max(optimal_k, 4)  # Minimum 4 clusters for microbiome data
optimal_k <- min(optimal_k, 15)  # Maximum 15 clusters to avoid overfitting

cat("Silhouette analysis results:\n")
cat("  Raw silhouette scores (k=2-20):", round(sil_scores, 3), "\n")
cat("  Penalized silhouette scores:", round(sil_scores_penalized, 3), "\n")
cat("  Optimal k (penalized):", optimal_k, "clusters\n")
cat("  Final optimal k (constrained):", optimal_k, "clusters\n")

# Cut the tree into the optimal number of clusters
clusters_numeric <- cutree(hc, k = optimal_k)
cat("Using", optimal_k, "clusters determined by adjusted silhouette analysis.\n")

# Step 10: Convert numeric cluster assignments to character cluster names
# Create dynamic cluster mapping based on optimal number of clusters
cluster_letters <- letters[1:min(optimal_k, 26)]  # Use letters a-z (max 26 clusters)
cluster_mapping <- setNames(cluster_letters, as.character(1:optimal_k))

# Convert numeric clusters to character clusters
clusters <- factor(cluster_mapping[as.character(clusters_numeric)], 
                  levels = cluster_letters)

# Define cluster colors (dynamic based on number of clusters)
# Use a predefined color palette that can handle up to 26 clusters
cluster_color_palette <- c(
  "#FF80DA", "#FF9289", "#F3AC00", "#89CC00", "#00DCA9", "#00D8EF", 
  "#9CB1FF", "#D69EFF", "#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4",
  "#FFEAA7", "#DDA0DD", "#98D8C8", "#F7DC6F", "#BB8FCE", "#85C1E9",
  "#F8C471", "#82E0AA", "#F1948A", "#85C1E9", "#D7BDE2", "#A9DFBF",
  "#F9E79F", "#AED6F1"
)
cluster_colors <- setNames(cluster_color_palette[1:optimal_k], cluster_letters)

# Step 11: Prepare data for visualization
mean_abundance_df <- as.data.frame(mean_abundance_scaled)
mean_abundance_df$genus <- rownames(mean_abundance_df)
mean_abundance_df$cluster <- clusters
mean_abundance_df$phylum <- tax_table(ubiome_filtered)[mean_abundance_df$genus, "Phylum"]

# Define phylum colors (STRICT DEFINITION - NEVER DEVIATE)
phylum_colors <- c(
  "Firmicutes"           = "#E69F00",  # orange
  "Bacteroidetes"        = "#56B4E9",  # blue
  "Actinobacteria"       = "#009E73",  # green
  "Proteobacteria"       = "#F0E442",  # yellow
  "Fusobacteria"         = "#CC6677",  # dark blue
  "Spirochaetae"         = "#D55E00",  # reddish-orange
  "Cyanobacteria"        = "#CC79A7",  # purple
  "Acidobacteria"        = "#999999",  # grey
  "Candidate division SR1" = "#AD7700",# brown
  "Planctomycetes"       = "#332288",  # deep indigo
  "Saccharibacteria"     = "#44AA99",  # teal
  "Synergistetes"        = "#88CCEE",  # light blue
  "Tenericutes"          = "#117733",  # dark green
  "unclassified"         = "#DDDDDD",  # very light grey
  "NA"                   = "#DDDDDD",  # very light grey
  "Verrucomicrobia"      = "#0072B2"   # reddish pink
)
mean_abundance_df$phylum <- factor(mean_abundance_df$phylum, levels = names(phylum_colors))

# Step 12: Enhanced parallel coordinate plot
p_parallel <- ggparcoord(
  data = mean_abundance_df,
  columns = 1:(ncol(mean_abundance_df) - 3),  # Exclude genus, cluster, phylum
  groupColumn = "phylum",
  scale = "globalminmax",
  title = "Genus Abundance Trends Across Age Groups by Cluster",
  alphaLines = 0.5,  # Reduce overplotting
  showPoints = FALSE, # Add points for clarity
  splineFactor = 20  # Smooth lines
) +
  facet_wrap(~ cluster, scales = "free_y", ncol = 2) +  # Free y-axis per cluster, 2 columns
  scale_color_manual(values = phylum_colors) +
  egg::theme_article() +  # White background, minimal grid
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 6),
    strip.text = element_text(size = 6, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  ) +
  labs(x = "Age Group", y = "Standardized Relative Abundance", color = "Phylum")
print(p_parallel)
cat("Generated enhanced parallel coordinate plot.\n")

# Step 13: PCA for dimensionality reduction
pca_result <- prcomp(mean_abundance_scaled, scale. = FALSE)  # Already scaled
pca_df <- as.data.frame(pca_result$x[, 1:2])  # First two PCs
pca_df$genus <- rownames(mean_abundance_scaled)
pca_df$cluster <- clusters
pca_df$phylum <- mean_abundance_df$phylum

# Calculate variance explained
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2) * 100
pc1_label <- sprintf("PC1 (%.1f%%)", var_explained[1])
pc2_label <- sprintf("PC2 (%.1f%%)", var_explained[2])

# PCA plot colored by cluster
p_pca_cluster <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = cluster_colors) +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "PCA of Genera by Cluster Assignment",
    x = pc1_label,
    y = pc2_label,
    color = "Cluster"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_pca_cluster)

# PCA plot colored by phylum
p_pca_phylum <- ggplot(pca_df, aes(x = PC1, y = PC2, color = phylum)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = phylum_colors) +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "PCA of Genera by Phylum",
    x = pc1_label,
    y = pc2_label,
    color = "Phylum"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_pca_phylum)

# Step 14: t-SNE for non-linear dimensionality reduction
set.seed(42)  # Set seed for t-SNE reproducibility
tsne_result <- Rtsne(mean_abundance_scaled, dims = 2, perplexity = min(30, (nrow(mean_abundance_scaled) - 1) / 3), verbose = TRUE)
tsne_df <- as.data.frame(tsne_result$Y)
colnames(tsne_df) <- c("Dim1", "Dim2")
tsne_df$genus <- rownames(mean_abundance_scaled)
tsne_df$cluster <- clusters
tsne_df$phylum <- mean_abundance_df$phylum

# t-SNE plot colored by cluster
p_tsne_cluster <- ggplot(tsne_df, aes(x = Dim1, y = Dim2, color = cluster)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = cluster_colors) +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "t-SNE of Genera by Cluster Assignment",
    x = "t-SNE Dimension 1",
    y = "t-SNE Dimension 2",
    color = "Cluster"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_tsne_cluster)

# t-SNE plot colored by phylum
p_tsne_phylum <- ggplot(tsne_df, aes(x = Dim1, y = Dim2, color = phylum)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = phylum_colors) +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "t-SNE of Genera by Phylum",
    x = "t-SNE Dimension 1",
    y = "t-SNE Dimension 2",
    color = "Phylum"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_tsne_phylum)

# Step 15: Create meaningful labels for heatmap rows
# Extract taxonomy table from phyloseq object and add OTU as a column
tax_df <- as.data.frame(tax_table(ubiome_filtered))
tax_df$OTU <- rownames(tax_df)  # Convert row names (OTU IDs) to a column

# Join taxonomy data with mean_abundance_df and create labels
mean_abundance_df <- mean_abundance_df %>%
  left_join(tax_df, by = c("genus" = "OTU")) %>%
  # Clean the OTU ID by removing "RSV_" and "_relative"
  mutate(cleaned_otu = gsub("^RSV_|^OTU_", "", genus)) %>%
  mutate(cleaned_otu = gsub("_relative$", "", cleaned_otu)) %>%
  # Create labels: use Genus if valid, otherwise "NA; <cleaned_otu>"
  mutate(label = if_else(
    is.na(Genus) | tolower(Genus) == "unclassified",
    paste0("NA; ", cleaned_otu),
    as.character(Genus)
  )) %>%
  # Ensure unique labels by appending OTU ID if duplicates exist
  group_by(label) %>%
  mutate(label = if_else(n() > 1, paste0(label, "_", genus), label)) %>%
  ungroup()

# Update row names of the scaled abundance matrix with the new labels
rownames(mean_abundance_scaled) <- mean_abundance_df$label
# Step 16: Heatmap with cluster and phylum annotations
annotation_row <- data.frame(
  Cluster = clusters,
  Phylum = mean_abundance_df$phylum
)
rownames(annotation_row) <- mean_abundance_df$label  # Use labels instead of genus

pheatmap(
  mean_abundance_scaled,
  cluster_rows = hc,
  cluster_cols = FALSE,
  annotation_row = annotation_row,
  annotation_colors = list(Cluster = cluster_colors, Phylum = phylum_colors),
  show_rownames = TRUE,  # Display the new labels
  color = colorRampPalette(c("#053061", "#F7F7F7", "#67001F"))(100),
  main = "Heatmap of Standardized Genus Abundance by Age Group",
  fontsize = 6,
  border_color = NA,
  cellwidth = 10,  # Adjust cell width for better readability
  cellheight = 10  # Adjust cell height to fit labels
)
cat("Generated heatmap with cluster and phylum annotations.\n")

# Step 17: Cluster summary statistics
cluster_summary <- mean_abundance_df %>%
  group_by(cluster) %>%
  summarise(
    n_genera = n(),
    mean_abundance = mean(rowMeans(mean_abundance_scaled)),
    sd_abundance = sd(rowMeans(mean_abundance_scaled)),
    dominant_phyla = paste(names(sort(table(phylum), decreasing = TRUE))[1:2], collapse = ", ")
  )
print(cluster_summary)

# Step 18: Combine PCA and t-SNE plots for publication
combined_plot <- plot_grid(
  p_pca_cluster + theme(legend.position = "bottom"),
  p_pca_phylum + theme(legend.position = "bottom"),
  p_tsne_cluster + theme(legend.position = "bottom"),
  p_tsne_phylum + theme(legend.position = "bottom"),
  labels = c("A", "B", "C", "D"),
  ncol = 2,
  label_size = 14
)
print(combined_plot)

# Step 19: Save all plots and results using the PDF saving function
save_pdf_figure("15.3_parallel_coord_plot_faceted.pdf", width_mm = 100, height_mm = 90)
print(p_parallel)
dev.off()

save_pdf_figure("15.4_pca_cluster.pdf", width_mm = 44, height_mm = 30)
print(p_pca_cluster)
dev.off()

save_pdf_figure("15.5_pca_phylum.pdf", width_mm = 65, height_mm = 30)
print(p_pca_phylum)
dev.off()

save_pdf_figure("15.6_tsne_cluster.pdf", width_mm = 44, height_mm = 30)
print(p_tsne_cluster)
dev.off()

save_pdf_figure("15.7_tsne_phylum.pdf", width_mm = 65, height_mm = 30)
print(p_tsne_phylum)
dev.off()

save_pdf_figure("15.8_combined_pca_tsne.pdf", width_mm = 65, height_mm = 100)
print(combined_plot)
dev.off()

save_pdf_figure("15.9_heatmap.pdf", width_mm = 110, height_mm = 200)
pheatmap(
  mean_abundance_scaled,
  cluster_rows = hc,
  cluster_cols = FALSE,
  annotation_row = annotation_row,
  annotation_colors = list(Cluster = cluster_colors, Phylum = phylum_colors),
  show_rownames = TRUE,  # Displays the new Genus-based labels
  color = colorRampPalette(c("#053061", "#F7F7F7", "#67001F"))(100),
  main = "Heatmap of Standardized Genus Abundance by Age Group",
  fontsize = 6,
  border_color = NA,
  cellwidth = 5,
  cellheight = 5
)
dev.off()

# Save data frames as CSV
write.csv(mean_abundance_df, file.path(viz_out_path, "genus_cluster_assignments.csv"), row.names = FALSE)
write.csv(cluster_summary, file.path(viz_out_path, "cluster_summary.csv"), row.names = FALSE)
cat("Saved all plots and results.\n")
```

### Explanation of the Code

#### **Libraries**
- `phyloseq`: For microbiome data manipulation.
- `dplyr`: For data aggregation and summarization.
- `GGally`: For creating parallel coordinate plots via `ggparcoord`.

#### **Data Loading**
- The code assumes `ubiome_relative_updated.rds` is in the working directory. Adjust the file path as needed.

#### **Preprocessing**
- **Aggregation**: OTUs are aggregated to the genus level using `tax_glom`.
- **Filtering**: Genera with non-zero abundance in at least 10% of samples are retained using `filter_taxa`.
- **Age Grouping**: Ages (≥14) are binned into groups (14-19, 20-24, ..., 80-85) using `cut`. The last group includes ages 80-85 since the maximum age is 85.

#### **Mean Abundance Calculation**
- The OTU table is transposed, combined with age group metadata, and averaged per age group using `dplyr`.

#### **Differential Abundance**
- Kruskal-Wallis tests identify genera with significant abundance differences across age groups, with p-values adjusted for multiple testing.

#### **Clustering**
- The mean abundances of significant genera are standardized (mean = 0, SD = 1) per genus.
- Hierarchical clustering with Ward's D2 linkage is performed using `hclust`.
- **Cluster Naming**: Numeric cluster assignments are converted to character names (a, b, c, d, e, f, g, h) with specific color mappings.

#### **Visualization**
- A parallel coordinate plot visualizes standardized abundance trends across age groups, with lines colored by cluster assignment and phylum.
- PCA and t-SNE plots show cluster and phylum relationships.
- Heatmaps display abundance patterns with cluster and phylum annotations.

### Notes
- **Data Availability**: Ensure `ubiome_relative_updated.rds` is accessible.
- **Plot Customization**: The plot may be crowded if many genera are significant. Adjust `num_clusters` or add transparency (`alpha`) in `ggparcoord` if needed.
- **Cluster Colors**: Character cluster names (a-h) are mapped to specific colors that must never deviate.
- **Phylum Colors**: Strict color definitions for phylum categories that must never change.
- **Output**: The code prints progress messages and displays the plot. Uncomment the `ggsave` line to save it.

This code fully implements the requested analysis, adhering to the specified filtering and age-grouping requirements with the new cluster naming and coloring scheme.

Below is the updated R code starting from **Step 4: Compute mean relative abundance per age group** onwards, incorporating the requested modifications. The code assumes that steps 1–3 (loading libraries, aggregating to genus level, filtering genera, and defining age groups) have been executed correctly, and `ubiome_filtered` is available with the age groups defined in the sample data.

The updates include:
1. **Dynamic cluster selection**: Using the silhouette method to determine the optimal number of clusters in a data-driven way for hierarchical clustering.
2. **Faceted parallel coordinate plot**: Faceting the plot by cluster, with each panel representing one cluster in a single figure.
3. **Phylum-based coloring**: Coloring the lines in the parallel coordinate plot by phylum using the provided `phylum_colors` mapping.
4. **DTW and PAM clustering**: Adding an additional analysis using Dynamic Time Warping (DTW) distance with Partitioning Around Medoids (PAM) clustering, followed by a scatter plot visualization using PCA, with points colored by phylum.

Here is the complete updated code:



---

### Explanation of Enhancements

#### **Statistical Analysis**
- **Kruskal-Wallis Test**: Identifies differentially abundant genera across age groups with BH correction for multiple testing.
- **Clustering**: Hierarchical clustering with Ward's method, optimized using the silhouette method for robust cluster determination.
- **Cluster Summary**: Expanded to include the top two dominant phyla per cluster, providing biological context.

#### **Visualizations**
1. **Parallel Coordinate Plot**:
   - **Enhancements**: Added points (`showPoints = TRUE`), smoothed lines (`splineFactor = 10`), and transparency (`alphaLines = 0.5`) to reduce overplotting.
   - **Faceting**: Free y-axis per cluster for better pattern visibility, arranged in two columns.
   - **Aesthetics**: Minimal theme with bold titles, larger fonts, and a grid for readability.

2. **Dimensionality Reduction**:
   - **PCA**:
     - Shows genera in a 2D space, colored by cluster and phylum.
     - Includes variance explained in axis labels for interpretability.
     - Uses custom cluster colors and phylum colors.
   - **t-SNE**:
     - Non-linear reduction with adaptive perplexity for robustness.
     - Similar coloring and aesthetic improvements as PCA.

3. **Heatmap**:
   - Annotated with cluster and phylum information.
   - Custom color gradient for abundance levels, optimized for publication clarity.

4. **Combined Plot**:
   - PCA and t-SNE plots arranged in a 2x2 grid with labels (A-D) for a concise, publication-ready figure.

#### **Output**
- High-resolution PDF files for all plots.
- CSV exports for cluster assignments and summary statistics.

---

### Adapting the Paper's Figures to Your Dataset

The paper's Figures 1 and 2 analyze fecal microbiome data across age and populations. Since your dataset is from the oral microbiome of a single population spanning ages 14 to 85, we'll modify the approach to explore how microbial communities evolve with age within this range. Here's the plan:

#### **Figure 1: Microbial Community Composition and Age**
The original Figure 1 has three panels:
- **Panel a**: UniFrac distances between children and adults decrease with age.
- **Panel b**: Interpersonal variation is higher in children than adults across populations.
- **Panel c**: PCoA shows clustering by population.

**Adaptation**:
- **Panel a**: We'll calculate the average UniFrac distance from each individual to a reference group (e.g., older adults aged 60-85) and plot it against age to see if microbial communities stabilize with age.
- **Panel b**: We'll compute pairwise UniFrac distances within age groups (e.g., 14-19, 20-24, etc.) to assess interpersonal variation across ages.
- **Panel c**: We'll use PCoA to visualize how microbial communities cluster by age groups, replacing population with age as the variable of interest.

#### **Figure 2: Bacterial Diversity Increases with Age**
The original Figure 2 has three panels:
- **Panels a-c**: Observed OTUs increase with age across populations.

**Adaptation**:
- **Panel a**: Plot Observed OTUs against age for all individuals (14-85) to identify trends.
- **Panel b**: Create boxplots of Observed OTUs by age group to compare diversity across age ranges.

Since your data starts at age 14, we won't focus on early childhood (unlike the paper's 0-3 years), but we'll capture trends across adolescence to old age.

---

### R Code to Generate the Plots

### Complete R Code for Additional Analysis

```{r}
# Set seed for reproducibility
set.seed(42)

# Load required libraries
library(dtw)          # For DTW distance computation
library(cluster)      # For PAM clustering and silhouette method
library(ggplot2)      # For PCA visualization

# Assuming mean_abundance_scaled is available
# This should be a matrix with genera as rows and age groups as columns, already standardized
# Example: mean_abundance_scaled <- scale(mean_abundance_data)

# Step 1: Compute DTW distance matrix
n <- nrow(mean_abundance_scaled)
dtw_matrix <- matrix(0, n, n)  # Initialize empty distance matrix
for (i in 1:(n-1)) {
  for (j in (i+1):n) {
    # Compute DTW distance between each pair of genera
    dtw_matrix[i, j] <- dtw(mean_abundance_scaled[i,], mean_abundance_scaled[j,])$distance
    dtw_matrix[j, i] <- dtw_matrix[i, j]  # Symmetric matrix
  }
}
cat("Computed DTW distance matrix.\n")

# Convert the DTW matrix to a dist object for clustering
dtw_dist <- as.dist(dtw_matrix)

# Step 2: Determine optimal number of clusters using silhouette method
sil_scores_dtw <- numeric(9)  # Store silhouette scores for k = 2 to 10
for (k in 2:10) {
  # Perform PAM clustering for each k
  pam_result <- pam(dtw_dist, k = k)
  # Compute silhouette widths
  sil <- silhouette(pam_result$clustering, dtw_dist)
  sil_scores_dtw[k - 1] <- mean(sil[, "sil_width"])  # Average silhouette width
}
# Find the optimal number of clusters
optimal_k_dtw <- which.max(sil_scores_dtw) + 1
cat("Optimal number of clusters for PAM with DTW:", optimal_k_dtw, "\n")

# Step 3: Perform PAM clustering with the optimal number of clusters
pam_result <- pam(dtw_dist, k = optimal_k_dtw)
clusters_dtw <- pam_result$clustering  # Cluster assignments for each genus

# Step 4: Perform PCA for visualization
pca_result <- prcomp(mean_abundance_scaled, scale. = FALSE)  # Data is already scaled
pca_df <- as.data.frame(pca_result$x[, 1:2])  # Extract first two principal components
pca_df$genus <- rownames(mean_abundance_scaled)  # Add genus names
pca_df$cluster <- factor(clusters_dtw)  # Add cluster assignments

# Step 5: Add phylum information (optional, uncomment and adjust if available)
# If taxonomic information is available (e.g., from a phyloseq object):
# pca_df$phylum <- tax_table(ubiome_filtered)[pca_df$genus, "Phylum"]
# For now, we'll proceed without phylum data unless provided

# Define phylum colors (optional, used if phylum data is available)
phylum_colors <- c(
  "Firmicutes"           = "#E69F00",  # orange
  "Bacteroidetes"        = "#56B4E9",  # blue
  "Actinobacteria"       = "#009E73",  # green
  "Proteobacteria"       = "#F0E442",  # yellow
  "Fusobacteria"         = "#CC6677",  # dark blue
  "Spirochaetae"         = "#D55E00",  # reddish-orange
  "Cyanobacteria"        = "#CC79A7",  # purple
  "Acidobacteria"        = "#999999",  # grey
  "Candidate division SR1" = "#AD7700",# brown
  "Planctomycetes"       = "#332288",  # deep indigo
  "Saccharibacteria"     = "#44AA99",  # teal
  "Synergistetes"        = "#88CCEE",  # light blue
  "Tenericutes"          = "#117733",  # dark green
  "unclassified"         = "#DDDDDD",  # very light grey
  "Verrucomicrobia"      = "#0072B2"   # reddish pink
)

# Step 6: Visualize results with PCA
# Plot 1: Colored by cluster
p_pca_cluster <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "PCA of Genera Clusters (PAM with DTW) - Colored by Cluster",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Cluster"
  )
print(p_pca_cluster)

# Plot 2: Colored by phylum (uncomment if phylum data is added)
# p_pca_phylum <- ggplot(pca_df, aes(x = PC1, y = PC2, color = phylum)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = phylum_colors) +
#   theme_minimal() +
#   labs(
#     title = "PCA of Genera Clusters (PAM with DTW) - Colored by Phylum",
#     x = "Principal Component 1",
#     y = "Principal Component 2",
#     color = "Phylum"
#   )
# print(p_pca_phylum)

# Optional: Save the plots
# ggsave("pca_dtw_pam_cluster.png", p_pca_cluster, width = 8, height = 6, dpi = 300)
# ggsave("pca_dtw_pam_phylum.png", p_pca_phylum, width = 8, height = 6, dpi = 300)
```

---

### Explanation of the Code

#### **1. Loading Libraries**
- **`dtw`**: Computes Dynamic Time Warping distances between time series (e.g., abundance across age groups).
- **`cluster`**: Provides PAM clustering and silhouette analysis functions.
- **`ggplot2`**: Creates visualizations of the PCA results.

#### **2. DTW Distance Matrix**
- A nested loop calculates the DTW distance between each pair of genera based on their abundance profiles across age groups.
- The result is a symmetric matrix (`dtw_matrix`), which is converted to a `dist` object (`dtw_dist`) for clustering.

#### **3. Optimal Number of Clusters**
- The silhouette method tests PAM clustering for `k = 2` to `10`.
- The `k` with the highest average silhouette width is selected as the optimal number of clusters (`optimal_k_dtw`).

#### **4. PAM Clustering**
- PAM clustering is applied to the DTW distance matrix using the optimal `k`.
- Each genus is assigned to a cluster, stored in `clusters_dtw`.

#### **5. PCA for Visualization**
- PCA reduces the dimensionality of `mean_abundance_scaled` to two principal components for plotting.
- A data frame (`pca_df`) is created with PC1, PC2, genus names, and cluster assignments.
- Phylum information can be added if available (e.g., from a taxonomic table).

#### **6. Visualization**
- **Cluster Plot**: Shows genera in PCA space, colored by their PAM cluster assignments.
- **Phylum Plot** (optional): Colors points by phylum, using a predefined color scheme (`phylum_colors`). Uncomment this section if phylum data is integrated.

#### **7. Saving Plots** (optional)
- Uncomment the `ggsave` lines to save the plots as PNG files.

---

### Notes
- **Data Assumption**: The code assumes `mean_abundance_scaled` is a pre-existing, standardized matrix. If this is not available, you'll need to generate it beforehand (e.g., using `scale()` on raw abundance data).
- **Phylum Information**: The phylum-colored plot is commented out because it requires taxonomic data (e.g., from a `tax_table`). If you have this data, uncomment and adjust the relevant lines.
- **Performance**: The DTW distance computation uses a double loop, which may be slow for large datasets. For optimization, consider parallel processing (e.g., with the `parallel` package) if needed.

This code is fully standalone and executable, addressing the error in the user's query (`object 'dtw_dist' not found`) by ensuring `dtw_dist` is properly defined before use.
