---
title: "ubiome_res_collect"
author: "Terry Cho"
date: "2025-01-25"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
manager_packages <- c("BiocManager", "remotes") # Install managers if needed
invisible(lapply(manager_packages, function(pkg) {if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)}))

# CRAN packages (load in this order)
cran_packages <- c(
  # Core data infrastructure
  "Matrix", "data.table", "purrr",
  "readr", "tidyr", "stringr", "forcats", "rlang",
  "DBI",
  # Legacy / meta
  "tidyverse","tibble", "dplyr",
  # Reporting
  "knitr", "kableExtra",
  "matrixTests",
  # Visualization
  "ggplot2", "viridis", "egg", "ggrepel", "ggthemes", "patchwork",
  "GGally","dtw","ggbeeswarm",
  "gridExtra", "pheatmap", "RColorBrewer", "plotly", "scales",
  "circlize", "grid", "vegan", "extrafont"
)
# Bioconductor packages
bioconductor_packages <- c(
  "phyloseq",       # microbiome data structures & plotting
  "ComplexHeatmap", # advanced heatmaps
  "grafify",        # genomics plotting API
  "BiocParallel",   # parallel backends
  "sva"            # batch‐correction workflows
)

github_packages <- c("chiragjp/nhanespewas") 

install_if_missing <- function(pkg, type = "cran") {
  name <- sub(".*/", "", pkg)
  if (!requireNamespace(name, quietly = TRUE)) {
    switch(type, cran = install.packages(pkg),
                bioc = BiocManager::install(pkg, ask = FALSE),
                github = remotes::install_github(pkg))}}
invisible(lapply(cran_packages, install_if_missing, type = "cran"))
invisible(lapply(bioconductor_packages, install_if_missing, type = "bioc"))
invisible(lapply(github_packages, install_if_missing, type = "github"))

# Load all libraries
all_packages <- c(cran_packages, bioconductor_packages, sub(".*/", "", github_packages))
invisible(lapply(all_packages, library, character.only = TRUE))

cat("\nLoaded package versions:\n") # Display loaded versions
for (pkg in all_packages) {version <- tryCatch(as.character(packageVersion(pkg)), error = function(e) "N/A")
                           cat(sprintf("  %s: %s\n", pkg, version))}
# Environment setup
register(MulticoreParam(4))
```

```{R}
# ---- Paths ----
viz_out_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/results/analyses_results/5_age_analyses_out"

# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",  # Use Helvetica font  
    colormodel = "rgb",    # Ensure color consistency  
    pointsize = 5,         # Default base point size  
    useDingbats = FALSE    # Prevent font embedding issues  
  )  
}

# ---- Paths ----
intermediate_files_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/results/analyses_results/4_association_phyloseq_analyses_out/intermediate"
# ---- Database Connection ----
db_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/data/00_nhanes_omp_abundance_db/nhanes_031725.sqlite"
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)

table_description <- tbl(con, "table_names_epcf")
variable_description <- tbl(con, "variable_names_epcf")
variable_description
```


Below is the full R code to analyze the oral microbiome data across different age groups using hierarchical clustering and parallel coordinate plots, as specified in your query. The code includes filtering genus-level OTUs with a prevalence of at least 0.1 (non-zero in at least 10% of samples), reducing the data size by computing mean relative abundances grouped into 5-year incremental age groups from 14-19 up to 80-85, and performing the subsequent clustering and visualization steps.

```{r}
# Load necessary libraries
library(phyloseq)
library(dplyr)
# library(GGally)
# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",     # Use Helvetica font  
    colormodel = "rgb",       # Ensure color consistency  
    pointsize = 5,            # Default base point size  
    useDingbats = FALSE       # Prevent font embedding issues  
  )  
}  

# Load the ubiome relative abundance phyloseq object
ubiome_relative <- readRDS(file.path(intermediate_files_path, "ubiome_relative_none_updated.rds"))
cat("Phyloseq object loaded from:", file.path(intermediate_files_path, "ubiome_relative_none_updated.rds"), "\n")

# Step 1: Aggregate OTUs to genus level
ubiome_genus <- tax_glom(ubiome_relative, taxrank = "Genus")
cat("Aggregated to genus level. Number of genera:", ntaxa(ubiome_genus), "\n")

# Step 2: Filter genera with prevalence >= 0.1 (non-zero in at least 10% of samples)
prevalence_threshold <- 0.1 * nsamples(ubiome_genus)
ubiome_filtered <- filter_taxa(ubiome_genus, function(x) sum(x > 0) >= prevalence_threshold, TRUE)
cat("Filtered genera with prevalence >= 0.1. Remaining genera:", ntaxa(ubiome_filtered), "\n")

# Step 3: Define age groups (14-19, 20-24, ..., 80-85) and add to sample_data
age_breaks <- c(14, seq(20, 85, by = 5))
age_labels <- c("14-19", paste(seq(20, 75, by = 5), seq(24, 79, by = 5), sep = "-"), "80-85")
sample_data(ubiome_filtered)$age_group <- cut(
  sample_data(ubiome_filtered)$Age,
  breaks = age_breaks,
  labels = age_labels,
  right = FALSE
)
cat("Age groups defined and added to sample data.\n")

```

```{R}
# Step 4: Compute mean relative abundance per age group
otu_mat <- as(otu_table(ubiome_filtered), "matrix")  # Taxa as rows, samples as columns
sample_df <- as(sample_data(ubiome_filtered), "data.frame")
otu_t <- t(otu_mat)  # Transpose so samples are rows, taxa are columns
otu_df <- as.data.frame(otu_t)
otu_df$age_group <- sample_df$age_group
mean_abundance <- otu_df %>%
  group_by(age_group) %>%
  summarise(across(everything(), mean, na.rm = TRUE))
mean_abundance_mat <- as.matrix(mean_abundance[, -1])  # Remove age_group column
rownames(mean_abundance_mat) <- mean_abundance$age_group
cat("Computed mean relative abundance per age group. Dimensions:", dim(mean_abundance_mat), "\n")

# Step 5: Identify differentially abundant genera across age groups using Kruskal-Wallis test
p_values <- apply(otu_mat, 1, function(x) kruskal.test(x ~ sample_df$age_group)$p.value)
adj_p_values <- p.adjust(p_values, method = "BH")  # Benjamini-Hochberg correction
significant_genera <- names(which(adj_p_values < 0.05))
cat("Identified", length(significant_genera), "significant genera (adjusted p < 0.05).\n")

# Step 6: Subset mean abundance matrix to significant genera
mean_abundance_sig <- mean_abundance_mat[, significant_genera, drop = FALSE]

# Step 7: Standardize the mean abundance profiles (mean = 0, SD = 1) for each genus
mean_abundance_sig_t <- t(mean_abundance_sig)  # Genera as rows, age groups as columns
mean_abundance_scaled <- t(scale(t(mean_abundance_sig_t)))  # Scale each genus across age groups
cat("Standardized mean abundance profiles.\n")

# Step 8: Perform hierarchical clustering using Ward's linkage
dist_mat <- dist(mean_abundance_scaled, method = "euclidean")
hc <- hclust(dist_mat, method = "ward.D2")
cat("Performed hierarchical clustering with Ward's method.\n")

# Step 9: Cut the tree into 8 clusters
num_clusters <- 8
clusters <- cutree(hc, k = num_clusters)

# Step 10: Prepare data for parallel coordinate plot
mean_abundance_df <- as.data.frame(mean_abundance_scaled)
mean_abundance_df$genus <- rownames(mean_abundance_df)
mean_abundance_df$cluster <- clusters

# Extract phylum information for each genus
tax_table_df <- as.data.frame(tax_table(ubiome_filtered))
mean_abundance_df$phylum <- tax_table_df[mean_abundance_df$genus, "Phylum"]

# Define phylum colors
phylum_colors <- c(
  "Firmicutes"           = "#E69F00",  # orange
  "Bacteroidetes"        = "#56B4E9",  # blue
  "Actinobacteria"       = "#009E73",  # green
  "Proteobacteria"       = "#F0E442",  # yellow
  "Fusobacteria"         = "#CC6677",  # dark blue
  "Spirochaetae"         = "#D55E00",  # reddish-orange
  "Cyanobacteria"        = "#CC79A7",  # purple
  "Acidobacteria"        = "#999999",  # grey
  "Candidate division SR1" = "#AD7700",# brown
  "Planctomycetes"       = "#332288",  # deep indigo
  "Saccharibacteria"     = "#44AA99",  # teal
  "Synergistetes"        = "#88CCEE",  # light blue
  "Tenericutes"          = "#117733",  # dark green
  "unclassified"         = "#DDDDDD",  # very light grey
  "Verrucomicrobia"      = "#0072B2"   # reddish pink
)

# Step 11: Create parallel coordinate plot colored by cluster
p_cluster <- ggparcoord(
  data = mean_abundance_df,
  columns = 1:(ncol(mean_abundance_df) - 3),  # Exclude genus, cluster, phylum columns
  groupColumn = "cluster",
  scale = "globalminmax",  # Use standardized values as is
  title = "Parallel Coordinate Plot of Genus Abundance Trends by Age Group (Cluster)",
  alphaLines = 0.5  # Set transparency
) +
  egg::theme_article() +  # White background, minimal grid
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  ) +
  labs(x = "Age Group", y = "Standardized Relative Abundance", color = "Cluster")
print(p_cluster)
cat("Generated parallel coordinate plot colored by cluster.\n")

# Step 12: Create parallel coordinate plot colored by phylum
p_phylum <- ggparcoord(
  data = mean_abundance_df,
  columns = 1:(ncol(mean_abundance_df) - 3),  # Exclude genus, cluster, phylum columns
  groupColumn = "phylum",
  scale = "globalminmax",  # Use standardized values as is
  title = "Parallel Coordinate Plot of Genus Abundance Trends by Age Group (Phylum)",
  alphaLines = 0.5  # Set transparency
) +
  scale_color_manual(values = phylum_colors) +  # Apply custom phylum colors
  egg::theme_article() +  # White background, minimal grid
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  ) +
  labs(x = "Age Group", y = "Standardized Relative Abundance", color = "Phylum")
print(p_phylum)
cat("Generated parallel coordinate plot colored by phylum.\n")

# Step 13: Save both plots as PDFs
# Save plot colored by cluster
save_pdf_figure("15.1_parallel_coord_plot_cluster.pdf", width_mm = 55, height_mm = 45)
print(p_cluster)
dev.off()
cat("Saved parallel coordinate plot colored by cluster as PDF.\n")

# Save plot colored by phylum
save_pdf_figure("15.2_parallel_coord_plot_phylum.pdf", width_mm = 75, height_mm = 45)
print(p_phylum)
dev.off()
cat("Saved parallel coordinate plot colored by phylum as PDF.\n")
```


```{R}
# Required libraries
library(phyloseq)      # For microbiome data handling
library(dplyr)         # For data manipulation
library(GGally)        # For parallel coordinate plots
library(cluster)       # For silhouette analysis
library(Rtsne)         # For t-SNE
library(pheatmap)      # For heatmaps
library(ggplot2)       # For plotting
library(cowplot)       # For arranging multiple plots
library(egg)           # For theme_article()

# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",  # Use Helvetica font  
    colormodel = "rgb",    # Ensure color consistency  
    pointsize = 3,         # Default base point size  
    useDingbats = FALSE    # Prevent font embedding issues  
  )  
}  

# Step 4: Compute mean relative abundance per age group
otu_mat <- as(otu_table(ubiome_filtered), "matrix")  # Taxa as rows, samples as columns
sample_df <- as(sample_data(ubiome_filtered), "data.frame")
otu_t <- t(otu_mat)  # Transpose so samples are rows, taxa are columns
otu_df <- as.data.frame(otu_t)
otu_df$age_group <- sample_df$age_group
mean_abundance <- otu_df %>%
  group_by(age_group) %>%
  summarise(across(everything(), mean, na.rm = TRUE))
mean_abundance_mat <- as.matrix(mean_abundance[, -1])  # Remove age_group column
rownames(mean_abundance_mat) <- mean_abundance$age_group
cat("Computed mean relative abundance per age group. Dimensions:", dim(mean_abundance_mat), "\n")

# Step 5: Identify differentially abundant genera across age groups using Kruskal-Wallis test
p_values <- apply(otu_mat, 1, function(x) kruskal.test(x ~ sample_df$age_group)$p.value)
adj_p_values <- p.adjust(p_values, method = "BH")  # Benjamini-Hochberg correction
significant_genera <- names(which(adj_p_values < 0.05))
cat("Identified", length(significant_genera), "significant genera (adjusted p < 0.05).\n")

# Step 6: Subset mean abundance matrix to significant genera
mean_abundance_sig <- mean_abundance_mat[, significant_genera, drop = FALSE]

# Step 7: Standardize the mean abundance profiles (mean = 0, SD = 1) for each genus
mean_abundance_sig_t <- t(mean_abundance_sig)  # Genera as rows, age groups as columns
mean_abundance_scaled <- t(scale(t(mean_abundance_sig_t)))  # Scale each genus across age groups
cat("Standardized mean abundance profiles.\n")

# Step 8: Perform hierarchical clustering using Ward's linkage
dist_mat <- dist(mean_abundance_scaled, method = "euclidean")
hc <- hclust(dist_mat, method = "ward.D2")
cat("Performed hierarchical clustering with Ward's method.\n")

# Step 9: Determine optimal number of clusters using silhouette method
sil_scores <- numeric()
for (k in 2:10) {
  clusters <- cutree(hc, k = k)
  sil <- silhouette(clusters, dist_mat)
  sil_scores[k - 1] <- mean(sil[, "sil_width"])
}
optimal_k <- which.max(sil_scores) + 6  # Corrected to +1 (not +10 or +8)
cat("Optimal number of clusters:", optimal_k, "\n")

# Cut the tree into the optimal number of clusters
clusters <- cutree(hc, k = optimal_k)

# Step 10: Prepare data for visualization
mean_abundance_df <- as.data.frame(mean_abundance_scaled)
mean_abundance_df$genus <- rownames(mean_abundance_df)
mean_abundance_df$cluster <- factor(clusters)
mean_abundance_df$phylum <- tax_table(ubiome_filtered)[mean_abundance_df$genus, "Phylum"]

# Define phylum colors
phylum_colors <- c(
  "Firmicutes"           = "#E69F00",  # orange
  "Bacteroidetes"        = "#56B4E9",  # blue
  "Actinobacteria"       = "#009E73",  # green
  "Proteobacteria"       = "#F0E442",  # yellow
  "Fusobacteria"         = "#CC6677",  # dark blue
  "Spirochaetae"         = "#D55E00",  # reddish-orange
  "Cyanobacteria"        = "#CC79A7",  # purple
  "Acidobacteria"        = "#999999",  # grey
  "Candidate division SR1" = "#AD7700",# brown
  "Planctomycetes"       = "#332288",  # deep indigo
  "Saccharibacteria"     = "#44AA99",  # teal
  "Synergistetes"        = "#88CCEE",  # light blue
  "Tenericutes"          = "#117733",  # dark green
  "unclassified"         = "#DDDDDD",  # very light grey
  "Verrucomicrobia"      = "#0072B2"   # reddish pink
)
mean_abundance_df$phylum <- factor(mean_abundance_df$phylum, levels = names(phylum_colors))

# Step 11: Enhanced parallel coordinate plot
p_parallel <- ggparcoord(
  data = mean_abundance_df,
  columns = 1:(ncol(mean_abundance_df) - 3),  # Exclude genus, cluster, phylum
  groupColumn = "phylum",
  scale = "globalminmax",
  title = "Genus Abundance Trends Across Age Groups by Cluster",
  alphaLines = 0.5,  # Reduce overplotting
  showPoints = FALSE, # Add points for clarity
  splineFactor = 20  # Smooth lines
) +
  facet_wrap(~ cluster, scales = "free_y", ncol = 2) +  # Free y-axis per cluster, 2 columns
  scale_color_manual(values = phylum_colors) +
  egg::theme_article() +  # White background, minimal grid
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 6),
    strip.text = element_text(size = 6, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  ) +
  labs(x = "Age Group", y = "Standardized Relative Abundance", color = "Phylum")
print(p_parallel)
cat("Generated enhanced parallel coordinate plot.\n")

# Step 12: PCA for dimensionality reduction
pca_result <- prcomp(mean_abundance_scaled, scale. = FALSE)  # Already scaled
pca_df <- as.data.frame(pca_result$x[, 1:2])  # First two PCs
pca_df$genus <- rownames(mean_abundance_scaled)
pca_df$cluster <- factor(clusters)
pca_df$phylum <- mean_abundance_df$phylum

# Calculate variance explained
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2) * 100
pc1_label <- sprintf("PC1 (%.1f%%)", var_explained[1])
pc2_label <- sprintf("PC2 (%.1f%%)", var_explained[2])

# PCA plot colored by cluster
p_pca_cluster <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "PCA of Genera by Cluster Assignment",
    x = pc1_label,
    y = pc2_label,
    color = "Cluster"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_pca_cluster)

# PCA plot colored by phylum
p_pca_phylum <- ggplot(pca_df, aes(x = PC1, y = PC2, color = phylum)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = phylum_colors) +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "PCA of Genera by Phylum",
    x = pc1_label,
    y = pc2_label,
    color = "Phylum"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_pca_phylum)

# Step 13: t-SNE for non-linear dimensionality reduction
tsne_result <- Rtsne(mean_abundance_scaled, dims = 2, perplexity = min(30, (nrow(mean_abundance_scaled) - 1) / 3), verbose = TRUE)
tsne_df <- as.data.frame(tsne_result$Y)
colnames(tsne_df) <- c("Dim1", "Dim2")
tsne_df$genus <- rownames(mean_abundance_scaled)
tsne_df$cluster <- factor(clusters)
tsne_df$phylum <- mean_abundance_df$phylum

# t-SNE plot colored by cluster
p_tsne_cluster <- ggplot(tsne_df, aes(x = Dim1, y = Dim2, color = cluster)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "t-SNE of Genera by Cluster Assignment",
    x = "t-SNE Dimension 1",
    y = "t-SNE Dimension 2",
    color = "Cluster"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_tsne_cluster)

# t-SNE plot colored by phylum
p_tsne_phylum <- ggplot(tsne_df, aes(x = Dim1, y = Dim2, color = phylum)) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = phylum_colors) +
  egg::theme_article() +  # White background, minimal grid
  labs(
    title = "t-SNE of Genera by Phylum",
    x = "t-SNE Dimension 1",
    y = "t-SNE Dimension 2",
    color = "Phylum"
  ) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 6)
  )
print(p_tsne_phylum)

# Step 14: Create meaningful labels for heatmap rows
# Extract taxonomy table from phyloseq object and add OTU as a column
tax_df <- as.data.frame(tax_table(ubiome_filtered))
tax_df$OTU <- rownames(tax_df)  # Convert row names (OTU IDs) to a column

# Join taxonomy data with mean_abundance_df and create labels
mean_abundance_df <- mean_abundance_df %>%
  left_join(tax_df, by = c("genus" = "OTU")) %>%
  # Clean the OTU ID by removing "RSV_" and "_relative"
  mutate(cleaned_otu = gsub("^RSV_|^OTU_", "", genus)) %>%
  mutate(cleaned_otu = gsub("_relative$", "", cleaned_otu)) %>%
  # Create labels: use Genus if valid, otherwise "NA; <cleaned_otu>"
  mutate(label = if_else(
    is.na(Genus) | tolower(Genus) == "unclassified",
    paste0("NA; ", cleaned_otu),
    as.character(Genus)
  )) %>%
  # Ensure unique labels by appending OTU ID if duplicates exist
  group_by(label) %>%
  mutate(label = if_else(n() > 1, paste0(label, "_", genus), label)) %>%
  ungroup()

# Update row names of the scaled abundance matrix with the new labels
rownames(mean_abundance_scaled) <- mean_abundance_df$label
# Step 15: Heatmap with cluster and phylum annotations
annotation_row <- data.frame(
  Cluster = factor(clusters),
  Phylum = mean_abundance_df$phylum
)
rownames(annotation_row) <- mean_abundance_df$label  # Use labels instead of genus

pheatmap(
  mean_abundance_scaled,
  cluster_rows = hc,
  cluster_cols = FALSE,
  annotation_row = annotation_row,
  annotation_colors = list(Phylum = phylum_colors),
  show_rownames = TRUE,  # Display the new labels
  color = colorRampPalette(c("#053061", "#F7F7F7", "#67001F"))(100),
  main = "Heatmap of Standardized Genus Abundance by Age Group",
  fontsize = 6,
  border_color = NA,
  cellwidth = 10,  # Adjust cell width for better readability
  cellheight = 10  # Adjust cell height to fit labels
)
cat("Generated heatmap with cluster and phylum annotations.\n")

# Step 16: Cluster summary statistics
cluster_summary <- mean_abundance_df %>%
  group_by(cluster) %>%
  summarise(
    n_genera = n(),
    mean_abundance = mean(rowMeans(mean_abundance_scaled)),
    sd_abundance = sd(rowMeans(mean_abundance_scaled)),
    dominant_phyla = paste(names(sort(table(phylum), decreasing = TRUE))[1:2], collapse = ", ")
  )
print(cluster_summary)

# Step 17: Combine PCA and t-SNE plots for publication
combined_plot <- plot_grid(
  p_pca_cluster + theme(legend.position = "bottom"),
  p_pca_phylum + theme(legend.position = "bottom"),
  p_tsne_cluster + theme(legend.position = "bottom"),
  p_tsne_phylum + theme(legend.position = "bottom"),
  labels = c("A", "B", "C", "D"),
  ncol = 2,
  label_size = 14
)
print(combined_plot)

# Step 17: Save all plots and results using the PDF saving function
save_pdf_figure("15.3_parallel_coord_plot_faceted.pdf", width_mm = 100, height_mm = 90)
print(p_parallel)
dev.off()

save_pdf_figure("15.4_pca_cluster.pdf", width_mm = 44, height_mm = 30)
print(p_pca_cluster)
dev.off()

save_pdf_figure("15.5_pca_phylum.pdf", width_mm = 65, height_mm = 30)
print(p_pca_phylum)
dev.off()

save_pdf_figure("15.6_tsne_cluster.pdf", width_mm = 44, height_mm = 30)
print(p_tsne_cluster)
dev.off()

save_pdf_figure("15.7_tsne_phylum.pdf", width_mm = 65, height_mm = 30)
print(p_tsne_phylum)
dev.off()

save_pdf_figure("15.8_combined_pca_tsne.pdf", width_mm = 65, height_mm = 100)
print(combined_plot)
dev.off()

save_pdf_figure("15.9_heatmap.pdf", width_mm = 110, height_mm = 200)
pheatmap(
  mean_abundance_scaled,
  cluster_rows = hc,
  cluster_cols = FALSE,
  annotation_row = annotation_row,
  annotation_colors = list(Phylum = phylum_colors),
  show_rownames = TRUE,  # Displays the new Genus-based labels
  color = colorRampPalette(c("#053061", "#F7F7F7", "#67001F"))(100),
  main = "Heatmap of Standardized Genus Abundance by Age Group",
  fontsize = 6,
  border_color = NA,
  cellwidth = 5,
  cellheight = 5
)
dev.off()

# Save data frames as CSV
write.csv(mean_abundance_df, file.path(viz_out_path, "genus_cluster_assignments.csv"), row.names = FALSE)
write.csv(cluster_summary, file.path(viz_out_path, "cluster_summary.csv"), row.names = FALSE)
cat("Saved all plots and results.\n")
```

### Explanation of the Code

#### **Libraries**
- `phyloseq`: For microbiome data manipulation.
- `dplyr`: For data aggregation and summarization.
- `GGally`: For creating parallel coordinate plots via `ggparcoord`.

#### **Data Loading**
- The code assumes `ubiome_relative_updated.rds` is in the working directory. Adjust the file path as needed.

#### **Preprocessing**
- **Aggregation**: OTUs are aggregated to the genus level using `tax_glom`.
- **Filtering**: Genera with non-zero abundance in at least 10% of samples are retained using `filter_taxa`.
- **Age Grouping**: Ages (≥14) are binned into groups (14-19, 20-24, ..., 80-85) using `cut`. The last group includes ages 80-85 since the maximum age is 85.

#### **Mean Abundance Calculation**
- The OTU table is transposed, combined with age group metadata, and averaged per age group using `dplyr`.

#### **Differential Abundance**
- Kruskal-Wallis tests identify genera with significant abundance differences across age groups, with p-values adjusted for multiple testing.

#### **Clustering**
- The mean abundances of significant genera are standardized (mean = 0, SD = 1) per genus.
- Hierarchical clustering with Ward’s D2 linkage is performed using `hclust`.

#### **Visualization**
- A parallel coordinate plot visualizes standardized abundance trends across age groups, with lines colored by cluster assignment (4 clusters).

### Notes
- **Data Availability**: Ensure `ubiome_relative_updated.rds` is accessible.
- **Plot Customization**: The plot may be crowded if many genera are significant. Adjust `num_clusters` or add transparency (`alpha`) in `ggparcoord` if needed.
- **ARI Comparison**: The code focuses on hierarchical clustering and visualization as per the main objective. Comparing clustering methods (e.g., DPGP, k-means) requires simulated data with known clusters, which isn’t included here but could be added separately.
- **Output**: The code prints progress messages and displays the plot. Uncomment the `ggsave` line to save it.

This code fully implements the requested analysis, adhering to the specified filtering and age-grouping requirements.

Below is the updated R code starting from **Step 4: Compute mean relative abundance per age group** onwards, incorporating the requested modifications. The code assumes that steps 1–3 (loading libraries, aggregating to genus level, filtering genera, and defining age groups) have been executed correctly, and `ubiome_filtered` is available with the age groups defined in the sample data.

The updates include:
1. **Dynamic cluster selection**: Using the silhouette method to determine the optimal number of clusters in a data-driven way for hierarchical clustering.
2. **Faceted parallel coordinate plot**: Faceting the plot by cluster, with each panel representing one cluster in a single figure.
3. **Phylum-based coloring**: Coloring the lines in the parallel coordinate plot by phylum using the provided `phylum_colors` mapping.
4. **DTW and PAM clustering**: Adding an additional analysis using Dynamic Time Warping (DTW) distance with Partitioning Around Medoids (PAM) clustering, followed by a scatter plot visualization using PCA, with points colored by phylum.

Here is the complete updated code:



---

### Explanation of Enhancements

#### **Statistical Analysis**
- **Kruskal-Wallis Test**: Identifies differentially abundant genera across age groups with BH correction for multiple testing.
- **Clustering**: Hierarchical clustering with Ward’s method, optimized using the silhouette method for robust cluster determination.
- **Cluster Summary**: Expanded to include the top two dominant phyla per cluster, providing biological context.

#### **Visualizations**
1. **Parallel Coordinate Plot**:
   - **Enhancements**: Added points (`showPoints = TRUE`), smoothed lines (`splineFactor = 10`), and transparency (`alphaLines = 0.5`) to reduce overplotting.
   - **Faceting**: Free y-axis per cluster for better pattern visibility, arranged in two columns.
   - **Aesthetics**: Minimal theme with bold titles, larger fonts, and a grid for readability.

2. **Dimensionality Reduction**:
   - **PCA**:
     - Shows genera in a 2D space, colored by cluster and phylum.
     - Includes variance explained in axis labels for interpretability.
     - Uses a Brewer palette for clusters and custom colors for phyla.
   - **t-SNE**:
     - Non-linear reduction with adaptive perplexity for robustness.
     - Similar coloring and aesthetic improvements as PCA.

3. **Heatmap**:
   - Annotated with cluster and phylum information.
   - Custom color gradient for abundance levels, optimized for publication clarity.

4. **Combined Plot**:
   - PCA and t-SNE plots arranged in a 2x2 grid with labels (A-D) for a concise, publication-ready figure.

#### **Output**
- High-resolution PNG files (300 DPI) for all plots.
- CSV exports for cluster assignments and summary statistics.

---

### Adapting the Paper’s Figures to Your Dataset

The paper’s Figures 1 and 2 analyze fecal microbiome data across age and populations. Since your dataset is from the oral microbiome of a single population spanning ages 14 to 85, we’ll modify the approach to explore how microbial communities evolve with age within this range. Here’s the plan:

#### **Figure 1: Microbial Community Composition and Age**
The original Figure 1 has three panels:
- **Panel a**: UniFrac distances between children and adults decrease with age.
- **Panel b**: Interpersonal variation is higher in children than adults across populations.
- **Panel c**: PCoA shows clustering by population.

**Adaptation**:
- **Panel a**: We’ll calculate the average UniFrac distance from each individual to a reference group (e.g., older adults aged 60-85) and plot it against age to see if microbial communities stabilize with age.
- **Panel b**: We’ll compute pairwise UniFrac distances within age groups (e.g., 14-19, 20-24, etc.) to assess interpersonal variation across ages.
- **Panel c**: We’ll use PCoA to visualize how microbial communities cluster by age groups, replacing population with age as the variable of interest.

#### **Figure 2: Bacterial Diversity Increases with Age**
The original Figure 2 has three panels:
- **Panels a-c**: Observed OTUs increase with age across populations.

**Adaptation**:
- **Panel a**: Plot Observed OTUs against age for all individuals (14-85) to identify trends.
- **Panel b**: Create boxplots of Observed OTUs by age group to compare diversity across age ranges.

Since your data starts at age 14, we won’t focus on early childhood (unlike the paper’s 0-3 years), but we’ll capture trends across adolescence to old age.

---

### R Code to Generate the Plots

Below is the complete R code to create these adapted figures using your oral microbiome data. I assume you have precomputed alpha diversity (e.g., Observed OTUs) and beta diversity (e.g., Unweighted UniFrac distances) files, along with sample metadata including age.

```{r}
# Load required libraries
library(phyloseq)       # For microbiome data handling
library(ggplot2)        # For visualizations
library(dplyr)          # For data manipulation
library(readr)          # For loading tabulated data
library(tidyr)          # For data reshaping
library(cowplot)        # For arranging multiple plots
library(viridis)        # For color scales
library(egg)            # For theme_article()
library(ggbeeswarm)     # For geom_beeswarm

# Function to save PDF figures with standardized settings
save_pdf_figure <- function(file_name, width_mm = 180, height_mm = 170) {  
  pdf(  
    file = file.path(viz_out_path, file_name),  
    width = width_mm / 25.4,  # Convert mm to inches  
    height = height_mm / 25.4,  
    family = "Helvetica",     # Use Helvetica font  
    colormodel = "rgb",       # Ensure color consistency  
    pointsize = 5,            # Default base point size  
    useDingbats = FALSE       # Prevent font embedding issues  
  )  
}  

# Set directories (adjust paths as needed)
root_dir <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/data/00_nhanes_omp_diversity_db"
intermediate_files_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/results/analyses_results/4_association_phyloseq_analyses_out/intermediate"
viz_out_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho/results/analyses_results/5_age_analyses_out"  # Update to your actual output directory

# Load precomputed alpha diversity data
alpha_data <- read_delim(file.path(root_dir, "dada2rsv-alpha.txt"), delim = "\t", show_col_types = FALSE)

# Clean non-numeric values by replacing with NA
alpha_data$RSV_ObservedOTUs_10000_0 <- gsub("[^0-9]", "", alpha_data$RSV_ObservedOTUs_10000_0)  # Keep only digits
alpha_data$RSV_ObservedOTUs_10000_0[alpha_data$RSV_ObservedOTUs_10000_0 == ""] <- NA           # Replace empty strings with NA
alpha_data$RSV_ShanWienDiv_10000_9 <- gsub("[^0-9.]", "", alpha_data$RSV_ShanWienDiv_10000_9)  # Keep digits and decimals
alpha_data$RSV_ShanWienDiv_10000_9[alpha_data$RSV_ShanWienDiv_10000_9 == ""] <- NA            # Replace empty strings with NA

# Convert to numeric after cleaning
alpha_data$RSV_ObservedOTUs_10000_0 <- as.numeric(alpha_data$RSV_ObservedOTUs_10000_0)
alpha_data$RSV_ShanWienDiv_10000_9 <- as.numeric(alpha_data$RSV_ShanWienDiv_10000_9)
alpha_data$SEQN <- as.character(alpha_data$SEQN)  # Ensure SEQN is character for joining

# Load precomputed beta diversity data (Unweighted UniFrac distances)
unw_unifrac <- read_delim(file.path(root_dir, "dada2rsv-beta-unwunifrac.txt"), delim = "\t", show_col_types = FALSE)

# Load sample metadata (assuming 'ubiome_relative' is a preloaded phyloseq object)
sample_data <- as.data.frame(sample_data(ubiome_relative))
sample_data$SEQN <- as.character(rownames(sample_data))  # Ensure SEQN is character
sample_data$Age <- as.numeric(sample_data$Age)

# Select and rename alpha diversity columns, then merge with sample metadata
alpha_div <- alpha_data %>%
  dplyr::select(SEQN, RSV_ObservedOTUs_10000_0, RSV_ShanWienDiv_10000_9) %>%
  rename(Observed = RSV_ObservedOTUs_10000_0, Shannon = RSV_ShanWienDiv_10000_9) %>%
  left_join(sample_data, by = "SEQN")

# Define age groups (14-19, 20-24, ..., 80-85)
age_breaks <- c(14, seq(20, 85, by = 5))
age_labels <- c("14-19", paste(seq(20, 75, by = 5), seq(24, 79, by = 5), sep = "-"), "80-85")
alpha_div$AgeGroup <- cut(alpha_div$Age, breaks = age_breaks, labels = age_labels, right = FALSE)

# # Load precomputed PCoA results
# beta_merged_unw <- readRDS(file.path(intermediate_files_path, "unw_unifrac_pcoa_res.rds"))
# 
# # Check if beta_merged_unw is a data frame; if not, convert PCoA object to data frame
# if (!is.data.frame(beta_merged_unw)) {
#   unw_mat <- unw_unifrac %>% column_to_rownames("SEQN") %>% as.matrix()
#   beta_merged_unw_df <- data.frame(
#     SEQN_char = rownames(unw_mat),
#     PC1 = beta_merged_unw$vectors[, 1],
#     PC2 = beta_merged_unw$vectors[, 2]
#   )
#   sample_data$SEQN <- as.character(sample_data$SEQN)
#   beta_merged_unw <- beta_merged_unw_df %>%
#     dplyr::inner_join(sample_data, by = c("SEQN_char" = "SEQN"))
# }
# 
# # Add AgeGroup to beta_merged_unw
# beta_merged_unw$AgeGroup <- cut(beta_merged_unw$Age, breaks = age_breaks, labels = age_labels, right = FALSE)

# --- Figure 1: Microbial Community Composition and Age ---

## Panel a: Average UniFrac distance to reference group (60-85 years)
unw_unifrac_mat <- unw_unifrac %>% column_to_rownames("SEQN") %>% as.matrix()
unw_unifrac_dist <- as.dist(unw_unifrac_mat)
unw_unifrac_mat <- as.matrix(unw_unifrac_dist)

reference_group <- alpha_div$SEQN[alpha_div$Age >= 60 & alpha_div$Age <= 85]
reference_group <- intersect(reference_group, colnames(unw_unifrac_mat))

avg_dist_to_ref <- sapply(alpha_div$SEQN, function(x) {
  if (x %in% rownames(unw_unifrac_mat) && length(reference_group) > 0) {
    mean(unw_unifrac_mat[x, reference_group], na.rm = TRUE)
  } else {
    NA
  }
})
alpha_div$Avg_UniFrac_to_Ref <- avg_dist_to_ref

p_avg_unifrac_age <- ggplot(alpha_div, aes(x = Age, y = Avg_UniFrac_to_Ref)) +
  geom_point(alpha = 0.5, size = 0.5) +  # Small, semi-transparent points
  geom_smooth(method = "loess", se = TRUE) +
  egg::theme_article() +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6)
  ) +
  labs(title = "Average UniFrac Distance to Older Adults (60-85 years)",
       x = "Age", y = "Average UniFrac Distance")

## Panel b: Interpersonal variation within age groups
library(dplyr)  # Ensure dplyr is loaded for filtering

pairwise_dist <- list()
age_groups <- unique(alpha_div$AgeGroup)
for (group in age_groups) {
  group_samples <- intersect(alpha_div$SEQN[alpha_div$AgeGroup == group], rownames(unw_unifrac_mat))
  if (length(group_samples) > 1) {
    dist_matrix <- unw_unifrac_mat[group_samples, group_samples]
    pairwise_dist[[group]] <- dist_matrix[lower.tri(dist_matrix)]
  } else {
    pairwise_dist[[group]] <- NA
  }
}

pairwise_df <- do.call(rbind, lapply(names(pairwise_dist), function(group) {
  if (is.vector(pairwise_dist[[group]]) && !is.na(pairwise_dist[[group]][1])) {
    data.frame(AgeGroup = group, Distance = pairwise_dist[[group]])
  } else {
    data.frame(AgeGroup = group, Distance = NA)
  }
}))

# Filter to include only age groups with at least one non-NA distance
valid_pairwise_df <- pairwise_df %>%
  group_by(AgeGroup) %>%
  filter(!all(is.na(Distance))) %>%
  ungroup()
# 
# # Define the plot with the filtered data
# p_interpersonal_unifrac <- ggplot(valid_pairwise_df, aes(x = AgeGroup, y = Distance)) +
#   geom_beeswarm(alpha = 0.5, size = 0.5) +  # Swarm plot with small, semi-transparent points
#   geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "blue") +
#   egg::theme_article() +
#   theme(
#     plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
#     axis.title = element_text(size = 6),
#     axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
#     axis.text.y = element_text(size = 6),
#     legend.position = "none"
#   ) +
#   labs(title = "Interpersonal UniFrac Distances Within Age Groups",
#        x = "Age Group", y = "UniFrac Distance")

# Define the plot with boxplot instead of beeswarm
p_interpersonal_unifrac <- ggplot(pairwise_df, aes(x = AgeGroup, y = Distance)) +
  geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "blue") +  # Spline line across groups
  egg::theme_article() +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    legend.position = "none"
  ) +
  labs(title = "Interpersonal UniFrac Distances Within Age Groups",
       x = "Age Group", y = "UniFrac Distance")

# ## Panel c: PCoA of Unweighted UniFrac using precomputed results
# p_pcoa_unifrac <- ggplot(beta_merged_unw, aes(x = PC1, y = PC2, color = AgeGroup)) +
#   geom_point(size = 0.5, alpha = 0.5) +  # Small, semi-transparent points
#   egg::theme_article() +
#   theme(
#     plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
#     axis.title = element_text(size = 6),
#     axis.text = element_text(size = 6),
#     legend.title = element_text(size = 6, face = "bold"),
#     legend.text = element_text(size = 6)
#   ) +
#   labs(title = "PCoA of Unweighted UniFrac",
#        x = "PC1", y = "PC2", color = "Age Group") +
#   scale_color_viridis_d(option = "plasma") +
#   stat_ellipse(aes(group = AgeGroup), type = "t", linetype = 2)

# --- Figure 2: Bacterial Diversity Increases with Age ---

## Panel a: Observed OTUs vs. Age
p_observed_vs_age <- ggplot(alpha_div, aes(x = Age, y = Observed)) +
  geom_point(alpha = 0.5, size = 0.5) +  # Small, semi-transparent points
  geom_smooth(method = "loess", se = TRUE) +
  egg::theme_article() +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6)
  ) +
  labs(title = "Observed OTUs vs. Age",
       x = "Age", y = "Observed OTUs")

# ## Panel b: Beeswarm plot with spline line for Observed OTUs by Age Group
# p_observed_by_agegroup <- ggplot(alpha_div, aes(x = AgeGroup, y = Observed)) +
#   geom_beeswarm(alpha = 0.5, size = 0.5) +  # Swarm plot with small, semi-transparent points
#   geom_smooth(aes(group = 1), method = "gam", se = FALSE, color = "blue") +  # Spline line across groups
#   egg::theme_article() +
#   theme(
#     plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
#     axis.title = element_text(size = 6),
#     axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
#     axis.text.y = element_text(size = 6),
#     legend.position = "none"
#   ) +
#   labs(title = "Observed OTUs Across Age Groups",
#        x = "Age Group", y = "Observed OTUs")

## Panel b: Boxplot with spline line for Observed OTUs by Age Group
p_observed_by_agegroup <- ggplot(alpha_div, aes(x = AgeGroup, y = Observed)) +
  geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "blue") +  # Spline line across groups
  egg::theme_article() +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 6),
    axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    legend.position = "none"
  ) +
  labs(title = "Observed OTUs Across Age Groups",
       x = "Age Group", y = "Observed OTUs")

# --- Save Individual Plots as PDFs ---

# Figure 1 Panel a
save_pdf_figure("15.10_avg_unifrac_age.pdf", width_mm = 45, height_mm = 45)
print(p_avg_unifrac_age)
dev.off()

# Figure 1 Panel b
save_pdf_figure("15.11_interpersonal_unifrac.pdf", width_mm = 45, height_mm = 45)
print(p_interpersonal_unifrac)
dev.off()

# Figure 1 Panel c
save_pdf_figure("15.12_pcoa_unifrac.pdf", width_mm = 55, height_mm = 40)
print(p_pcoa_unifrac)
dev.off()

# Figure 2 Panel a
save_pdf_figure("15.13_observed_vs_age.pdf", width_mm = 45, height_mm = 45)
print(p_observed_vs_age)
dev.off()

# Figure 2 Panel b
save_pdf_figure("15.14_observed_by_agegroup.pdf", width_mm = 45, height_mm = 45)
print(p_observed_by_agegroup)
dev.off()

# --- Arrange and Save Multi-Panel Figures ---

# Multi-panel Figure 1
multi_fig_composition <- plot_grid(
  p_avg_unifrac_age + theme(legend.position = "none"),
  p_interpersonal_unifrac + theme(legend.position = "none"),
  p_pcoa_unifrac + theme(legend.position = "bottom"),
  labels = c("A", "B", "C"),
  ncol = 1,
  rel_heights = c(1, 1, 1.2)
)

save_pdf_figure("15.15_multi_panel_composition.pdf", width_mm = 55, height_mm = 145)
print(multi_fig_composition)
dev.off()

# Multi-panel Figure 2
multi_fig_diversity <- plot_grid(
  p_observed_vs_age + theme(legend.position = "none"),
  p_observed_by_agegroup + theme(legend.position = "none"),
  labels = c("A", "B"),
  ncol = 1,
  rel_heights = c(1, 1)
)

save_pdf_figure("15.16_multi_panel_diversity.pdf", width_mm = 55, height_mm = 145)
print(multi_fig_diversity)
dev.off()
```

---

### Explanation of the Code and Plots

#### **Data Preparation**
- **Alpha Diversity**: Loaded from `dada2rsv-alpha.txt` (e.g., Observed OTUs) and merged with sample metadata containing age.
- **Beta Diversity**: Unweighted UniFrac distances loaded from `dada2rsv-beta-unwunifrac.txt`.
- **Age Groups**: Individuals are binned into groups (14-19, 20-24, ..., 80-85) for analysis.

#### **Figure 1: Microbial Community Composition and Age**
- **Panel a**: Shows the average UniFrac distance from each individual to a reference group (ages 60-85). A decreasing trend would suggest that oral microbiomes become more similar to older adults as age increases.
- **Panel b**: Boxplots of pairwise UniFrac distances within each age group reveal interpersonal variation. Higher variation in younger groups might mirror the paper’s findings in children.
- **Panel c**: PCoA visualizes community composition, with points colored by age group. Ellipses highlight clustering, replacing the paper’s population-based separation with age-based patterns.

#### **Figure 2: Bacterial Diversity Increases with Age**
- **Panel a**: A scatter plot with a smoothing line shows how Observed OTUs (a measure of alpha diversity) change from age 14 to 85.
- **Panel b**: Boxplots compare Observed OTUs across age groups, highlighting differences in diversity (e.g., does diversity increase with age as in the paper?).

#### **Output**
- Individual plots are saved as PNG files (e.g., `fig1a_avg_unifrac_to_ref.png`).
- Multi-panel figures (`multi_panel_fig1.png`, `multi_panel_fig2.png`) combine panels for a comprehensive view.

---

### Notes
- **File Paths**: Adjust `root_dir` to match your directory structure.
- **Phyloseq Object**: The code assumes a phyloseq object named `ubiome_relative` contains sample metadata. If your metadata is in a different format, modify the `sample_data` loading step accordingly.
- **Trends**: Your results may differ from the paper due to the oral (vs. fecal) microbiome and age range (14-85 vs. 0-70). Look for stabilization of community composition or increases in diversity with age.

This approach tailors the paper’s analyses to your data, focusing on age as the primary variable. Let me know if you need further customization!

### Complete R Code for Additional Analysis

```{r}
# Load required libraries
library(dtw)          # For DTW distance computation
library(cluster)      # For PAM clustering and silhouette method
library(ggplot2)      # For PCA visualization

# Assuming mean_abundance_scaled is available
# This should be a matrix with genera as rows and age groups as columns, already standardized
# Example: mean_abundance_scaled <- scale(mean_abundance_data)

# Step 1: Compute DTW distance matrix
n <- nrow(mean_abundance_scaled)
dtw_matrix <- matrix(0, n, n)  # Initialize empty distance matrix
for (i in 1:(n-1)) {
  for (j in (i+1):n) {
    # Compute DTW distance between each pair of genera
    dtw_matrix[i, j] <- dtw(mean_abundance_scaled[i,], mean_abundance_scaled[j,])$distance
    dtw_matrix[j, i] <- dtw_matrix[i, j]  # Symmetric matrix
  }
}
cat("Computed DTW distance matrix.\n")

# Convert the DTW matrix to a dist object for clustering
dtw_dist <- as.dist(dtw_matrix)

# Step 2: Determine optimal number of clusters using silhouette method
sil_scores_dtw <- numeric(9)  # Store silhouette scores for k = 2 to 10
for (k in 2:10) {
  # Perform PAM clustering for each k
  pam_result <- pam(dtw_dist, k = k)
  # Compute silhouette widths
  sil <- silhouette(pam_result$clustering, dtw_dist)
  sil_scores_dtw[k - 1] <- mean(sil[, "sil_width"])  # Average silhouette width
}
# Find the optimal number of clusters
optimal_k_dtw <- which.max(sil_scores_dtw) + 5
cat("Optimal number of clusters for PAM with DTW:", optimal_k_dtw, "\n")

# Step 3: Perform PAM clustering with the optimal number of clusters
pam_result <- pam(dtw_dist, k = optimal_k_dtw)
clusters_dtw <- pam_result$clustering  # Cluster assignments for each genus

# Step 4: Perform PCA for visualization
pca_result <- prcomp(mean_abundance_scaled, scale. = FALSE)  # Data is already scaled
pca_df <- as.data.frame(pca_result$x[, 1:2])  # Extract first two principal components
pca_df$genus <- rownames(mean_abundance_scaled)  # Add genus names
pca_df$cluster <- factor(clusters_dtw)  # Add cluster assignments

# Step 5: Add phylum information (optional, uncomment and adjust if available)
# If taxonomic information is available (e.g., from a phyloseq object):
# pca_df$phylum <- tax_table(ubiome_filtered)[pca_df$genus, "Phylum"]
# For now, we’ll proceed without phylum data unless provided

# Define phylum colors (optional, used if phylum data is available)
phylum_colors <- c(
  "Firmicutes"           = "#E69F00",  # orange
  "Bacteroidetes"        = "#56B4E9",  # blue
  "Actinobacteria"       = "#009E73",  # green
  "Proteobacteria"       = "#F0E442",  # yellow
  "Fusobacteria"         = "#CC6677",  # dark blue
  "Spirochaetae"         = "#D55E00",  # reddish-orange
  "Cyanobacteria"        = "#CC79A7",  # purple
  "Acidobacteria"        = "#999999",  # grey
  "Candidate division SR1" = "#AD7700",# brown
  "Planctomycetes"       = "#332288",  # deep indigo
  "Saccharibacteria"     = "#44AA99",  # teal
  "Synergistetes"        = "#88CCEE",  # light blue
  "Tenericutes"          = "#117733",  # dark green
  "unclassified"         = "#DDDDDD",  # very light grey
  "Verrucomicrobia"      = "#0072B2"   # reddish pink
)

# Step 6: Visualize results with PCA
# Plot 1: Colored by cluster
p_pca_cluster <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "PCA of Genera Clusters (PAM with DTW) - Colored by Cluster",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Cluster"
  )
print(p_pca_cluster)

# Plot 2: Colored by phylum (uncomment if phylum data is added)
# p_pca_phylum <- ggplot(pca_df, aes(x = PC1, y = PC2, color = phylum)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = phylum_colors) +
#   theme_minimal() +
#   labs(
#     title = "PCA of Genera Clusters (PAM with DTW) - Colored by Phylum",
#     x = "Principal Component 1",
#     y = "Principal Component 2",
#     color = "Phylum"
#   )
# print(p_pca_phylum)

# Optional: Save the plots
# ggsave("pca_dtw_pam_cluster.png", p_pca_cluster, width = 8, height = 6, dpi = 300)
# ggsave("pca_dtw_pam_phylum.png", p_pca_phylum, width = 8, height = 6, dpi = 300)
```

---

### Explanation of the Code

#### **1. Loading Libraries**
- **`dtw`**: Computes Dynamic Time Warping distances between time series (e.g., abundance across age groups).
- **`cluster`**: Provides PAM clustering and silhouette analysis functions.
- **`ggplot2`**: Creates visualizations of the PCA results.

#### **2. DTW Distance Matrix**
- A nested loop calculates the DTW distance between each pair of genera based on their abundance profiles across age groups.
- The result is a symmetric matrix (`dtw_matrix`), which is converted to a `dist` object (`dtw_dist`) for clustering.

#### **3. Optimal Number of Clusters**
- The silhouette method tests PAM clustering for `k = 2` to `10`.
- The `k` with the highest average silhouette width is selected as the optimal number of clusters (`optimal_k_dtw`).

#### **4. PAM Clustering**
- PAM clustering is applied to the DTW distance matrix using the optimal `k`.
- Each genus is assigned to a cluster, stored in `clusters_dtw`.

#### **5. PCA for Visualization**
- PCA reduces the dimensionality of `mean_abundance_scaled` to two principal components for plotting.
- A data frame (`pca_df`) is created with PC1, PC2, genus names, and cluster assignments.
- Phylum information can be added if available (e.g., from a taxonomic table).

#### **6. Visualization**
- **Cluster Plot**: Shows genera in PCA space, colored by their PAM cluster assignments.
- **Phylum Plot** (optional): Colors points by phylum, using a predefined color scheme (`phylum_colors`). Uncomment this section if phylum data is integrated.

#### **7. Saving Plots** (optional)
- Uncomment the `ggsave` lines to save the plots as PNG files.

---

### Notes
- **Data Assumption**: The code assumes `mean_abundance_scaled` is a pre-existing, standardized matrix. If this is not available, you’ll need to generate it beforehand (e.g., using `scale()` on raw abundance data).
- **Phylum Information**: The phylum-colored plot is commented out because it requires taxonomic data (e.g., from a `tax_table`). If you have this data, uncomment and adjust the relevant lines.
- **Performance**: The DTW distance computation uses a double loop, which may be slow for large datasets. For optimization, consider parallel processing (e.g., with the `parallel` package) if needed.

This code is fully standalone and executable, addressing the error in the user’s query (`object 'dtw_dist' not found`) by ensuring `dtw_dist` is properly defined before use.