---
title: "NHANES Oral Microbiome phyloseq Object Creation"
author: "Terry Cho"
date: "2025-01-19"
output: html_document
---

```{r knitr_init, echo=FALSE, warning=FALSE, cache=TRUE}
library(knitr)
library(rmdformats)
library(kableExtra)

# Global options
options(max.print = 75)
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  eval = TRUE,
  cache = FALSE,  # Disable cache for reproducibility with updated database
  echo = TRUE,
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(width = 85)
```

# NHANES Oral Microbiome phyloseq Object Creation Pipeline

> ** This pipeline creates phyloseq objects from the processed NHANES oral microbiome database**

Following the phyloseq tutorial: [https://vaulot.github.io/tutorials/Phyloseq_tutorial.html](https://vaulot.github.io/tutorials/Phyloseq_tutorial.html)


⚠️ **Prerequisites**
**Before running this script, ensure you have completed:**

1. **Chapter 0:** Data transformation (creates transformed microbiome tables)
2. **Chapter 2, Step 1:** Demographic processing (`process_demographics_complete.R`)
3. **Chapter 2, Step 2:** Database verification (`missing_tables_n_variables_diagnostic.R --processed`)


## **Setup & Libraries**

```{r libraries, message=FALSE}
# Set global chunk options for RMarkdown
knitr::opts_chunk$set(echo = TRUE)

# List of required packages
required_packages <- c(
  "phyloseq", "ape", "dplyr", "readr", "stringr", "data.tree", 
  "data.table", "tidyr", "readxl", "ggplot2", "tibble", 
  "tidyverse", "DBI", "RSQLite", "kableExtra", "compositions"
)

# Ensure CRAN mirror is set globally
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install missing packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(missing_packages) > 0) {
  install.packages(missing_packages, dependencies = TRUE)
}

# Load all required packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

## **Setup Paths and Database Connection**

```{r setup_paths}
base_path <- "/Users/byeongyeoncho/main/github/nhanes_oral_mirco_cho"

# Output paths
viz_out_path            <- file.path(base_path, "results/analyses_results/02_preprocess_db_n_phyloseq_out/figures_out")
intermediate_files_path <- file.path(base_path, "results/analyses_results/02_preprocess_db_n_phyloseq_out/intermediate")

# Create directories if they don't exist
dir.create(viz_out_path, recursive = TRUE, showWarnings = FALSE)
dir.create(intermediate_files_path, recursive = TRUE, showWarnings = FALSE)

# Database path - using PROCESSED database with comprehensive demographics
db_path <- file.path(base_path, "data/00_nhanes_omp_transformed_db/nhanes_oral_transformed_complete_processed.sqlite")

# Verify database exists
if (!file.exists(db_path)) {
  stop("❌ Processed database not found. Please run Chapter 2, Step 1 first:\n",
       "Rscript scripts/2_preprocess_db_n_phyloseq/process_demographics_complete.R")
}

# Connect to database
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)

cat("✅ Connected to processed database:", db_path, "\n")
```

## **Verify Database Status**

```{r verify_database}
# Check that we have the required transformed tables
required_tables <- c(
  # Count tables (for ubiome_counts)
  "DADA2RSV_GENUS_COUNT_F", "DADA2RSV_GENUS_COUNT_G",
  # Relative abundance tables (for ubiome_relative) 
  "DADA2RSV_GENUS_RELATIVE_F", "DADA2RSV_GENUS_RELATIVE_G",
  # Filtered + transformed relative abundance tables
  "DADA2RSV_GENUS_RELATIVE_F_none", "DADA2RSV_GENUS_RELATIVE_G_none",
  "DADA2RSV_GENUS_RELATIVE_F_clr", "DADA2RSV_GENUS_RELATIVE_G_clr", 
  "DADA2RSV_GENUS_RELATIVE_F_lognorm", "DADA2RSV_GENUS_RELATIVE_G_lognorm",
  # Processed demographic tables
  "DEMO_F", "DEMO_G"
)

available_tables <- dbListTables(con)
missing_tables <- setdiff(required_tables, available_tables)

if (length(missing_tables) > 0) {
  stop("❌ Missing required tables: ", paste(missing_tables, collapse = ", "), "\n",
       "Please ensure Chapter 0 (transformation) and Chapter 2 Step 1 (demographics) are complete.")
}

cat("✅ All required tables are available\n")

# Check demographic variables
demo_f_cols <- dbListFields(con, "DEMO_F")
demographic_indicators <- c("Gender", "AgeGroup", "EducationLevel", "Ethnicity", "PIRQuartile")
missing_demo_vars <- setdiff(demographic_indicators, demo_f_cols)

if (length(missing_demo_vars) > 0) {
  warning("⚠️ Missing demographic variables: ", paste(missing_demo_vars, collapse = ", "), "\n",
          "Consider running Chapter 2 Step 1 again.")
} else {
  cat("✅ Comprehensive demographic variables are available\n")
}
```

## **1. Create OTU Matrices for All Transformations**

### **1.1 Count Data (Original)**

```{r otu_counts}
cat("Creating count-based OTU matrix (reference structure)...\n")

# Read count tables
genus_count_f <- tbl(con, "DADA2RSV_GENUS_COUNT_F") %>% collect()
genus_count_g <- tbl(con, "DADA2RSV_GENUS_COUNT_G") %>% collect()

# Find overlapping SEQNs
overlapping_seqns_count <- intersect(genus_count_f$SEQN, genus_count_g$SEQN)
cat("Number of overlapping SEQNs in count data:", length(overlapping_seqns_count), "\n")

# Verify both count tables have identical column structure
if (!identical(colnames(genus_count_f), colnames(genus_count_g))) {
  stop("❌ Count tables F and G must have identical column structure")
}

# Get reference taxa columns and their base names (for mapping to transformed data)
reference_taxa_cols <- colnames(genus_count_f)[colnames(genus_count_f) != "SEQN"]
reference_base_names <- gsub("_count$", "", reference_taxa_cols)
cat("Reference taxa columns from count tables:", length(reference_taxa_cols), "\n")

# Combine count data
genus_count_combined <- bind_rows(genus_count_f, genus_count_g)

# Create count OTU matrix (using base names for consistent row names)
otu_mat_counts <- genus_count_combined %>%
  mutate(SEQN = as.character(SEQN)) %>%
  tidyr::pivot_longer(cols = matches("^RSV_genus.*_count$"), names_to = "otu", values_to = "Abundance") %>%
  # Convert to base names for consistent row naming across all matrices
  mutate(otu_base = gsub("_count$", "", otu)) %>%
  select(-otu) %>%
  tidyr::pivot_wider(names_from = SEQN, values_from = Abundance) %>%
  column_to_rownames("otu_base")

cat("✅ Count OTU matrix:", nrow(otu_mat_counts), "taxa ×", ncol(otu_mat_counts), "samples\n")
```

### **1.2 Transformed Relative Abundance Data**

```{r otu_matrices}
cat("Creating OTU matrices for 5 phyloseq objects...\n\n")

# Function to create OTU matrix from any table pair
create_otu_matrix_from_tables <- function(table_f_name, table_g_name, matrix_name, use_base_names = TRUE) {
  cat("Creating", matrix_name, "matrix from", table_f_name, "&", table_g_name, "...\n")
  
  # Read tables
  genus_f <- tbl(con, table_f_name) %>% collect()
  genus_g <- tbl(con, table_g_name) %>% collect()
  
  # Get taxa columns
  taxa_f <- colnames(genus_f)[colnames(genus_f) != "SEQN"]
  taxa_g <- colnames(genus_g)[colnames(genus_g) != "SEQN"]
  
  cat("  F table:", nrow(genus_f), "samples ×", length(taxa_f), "taxa\n")
  cat("  G table:", nrow(genus_g), "samples ×", length(taxa_g), "taxa\n")
  
  # Check if tables have identical structure
  if (!identical(colnames(genus_f), colnames(genus_g))) {
    # Take UNION of all taxa (not intersection) to preserve all data
    all_taxa <- union(taxa_f, taxa_g)
    cat("  Tables have different structures - expanding to", length(all_taxa), "total taxa (union)\n")
    
    # Add missing taxa to F table (fill with 0)
    missing_in_f <- setdiff(taxa_g, taxa_f)
    if (length(missing_in_f) > 0) {
      cat("  Adding", length(missing_in_f), "missing taxa to F table (filled with 0)\n")
      for (taxa in missing_in_f) {
        genus_f[[taxa]] <- 0
      }
    }
    
    # Add missing taxa to G table (fill with 0)
    missing_in_g <- setdiff(taxa_f, taxa_g)
    if (length(missing_in_g) > 0) {
      cat("  Adding", length(missing_in_g), "missing taxa to G table (filled with 0)\n")
      for (taxa in missing_in_g) {
        genus_g[[taxa]] <- 0
      }
    }
    
    # Reorder columns to match (SEQN first, then all taxa in consistent order)
    all_cols <- c("SEQN", all_taxa)
    genus_f <- genus_f %>% select(all_of(all_cols))
    genus_g <- genus_g %>% select(all_of(all_cols))
  }
  
  # Combine data
  genus_combined <- bind_rows(genus_f, genus_g)
  cat("  Combined:", nrow(genus_combined), "samples\n")
  
  # Create OTU matrix
  if (use_base_names) {
    # Use base names (remove suffix) for consistent row naming
    otu_mat <- genus_combined %>%
      mutate(SEQN = as.character(SEQN)) %>%
      tidyr::pivot_longer(cols = matches("^RSV_genus"), names_to = "otu", values_to = "Abundance") %>%
      mutate(otu_base = gsub("_(count|relative)$", "", otu)) %>%
      select(-otu) %>%
      tidyr::pivot_wider(names_from = SEQN, values_from = Abundance) %>%
      column_to_rownames("otu_base")
  } else {
    # Use original names
    otu_mat <- genus_combined %>%
      mutate(SEQN = as.character(SEQN)) %>%
      tidyr::pivot_longer(cols = matches("^RSV_genus"), names_to = "otu", values_to = "Abundance") %>%
      tidyr::pivot_wider(names_from = SEQN, values_from = Abundance) %>%
      column_to_rownames("otu")
  }
  
  cat("  ✅", matrix_name, "matrix:", nrow(otu_mat), "taxa ×", ncol(otu_mat), "samples\n\n")
  return(otu_mat)
}

# 1. Create ubiome_counts matrix (raw counts, no filtering)
otu_mat_counts <- create_otu_matrix_from_tables(
  "DADA2RSV_GENUS_COUNT_F", 
  "DADA2RSV_GENUS_COUNT_G", 
  "ubiome_counts"
)

# 2. Create ubiome_relative matrix (relative abundance, no filtering)
otu_mat_relative <- create_otu_matrix_from_tables(
  "DADA2RSV_GENUS_RELATIVE_F", 
  "DADA2RSV_GENUS_RELATIVE_G", 
  "ubiome_relative"
)

# 3. Create ubiome_relative_none matrix (0.1% prevalence filtered, no transformation)
otu_mat_relative_none <- create_otu_matrix_from_tables(
  "DADA2RSV_GENUS_RELATIVE_F_none", 
  "DADA2RSV_GENUS_RELATIVE_G_none", 
  "ubiome_relative_none"
)

# 4. Create ubiome_relative_clr matrix (0.1% prevalence filtered, CLR transformed)
otu_mat_relative_clr <- create_otu_matrix_from_tables(
  "DADA2RSV_GENUS_RELATIVE_F_clr", 
  "DADA2RSV_GENUS_RELATIVE_G_clr", 
  "ubiome_relative_clr"
)

# 5. Create ubiome_relative_lognorm matrix (0.1% prevalence filtered, log-normal transformed)
otu_mat_relative_lognorm <- create_otu_matrix_from_tables(
  "DADA2RSV_GENUS_RELATIVE_F_lognorm", 
  "DADA2RSV_GENUS_RELATIVE_G_lognorm", 
  "ubiome_relative_lognorm"
)

# Summary of all matrices
cat("📏 Summary of all OTU matrices:\n")
matrices <- list(
  "ubiome_counts" = otu_mat_counts,
  "ubiome_relative" = otu_mat_relative,
  "ubiome_relative_none" = otu_mat_relative_none,
  "ubiome_relative_clr" = otu_mat_relative_clr,
  "ubiome_relative_lognorm" = otu_mat_relative_lognorm
)

for (name in names(matrices)) {
  mat <- matrices[[name]]
  cat("  ", name, ":", nrow(mat), "taxa ×", ncol(mat), "samples\n")
}
```

## **2. Create Taxonomy Table**

```{r taxonomy_table}
cat("Creating taxonomy table...\n")

# Get taxonomy information from metadata
# First, let's check what taxa names we need to match
otu_taxa_names <- rownames(otu_mat_counts)
cat("Sample OTU names:", head(otu_taxa_names, 3), "\n")

# Get all available variable names to understand the structure
available_vars <- tbl(con, "variable_names_epcf") %>%
  collect() %>%
  filter(Data.File.Name == "DADA2RSV_GENUS_COUNT_G")

cat("Available variables in metadata:", nrow(available_vars), "\n")
cat("Sample variable names:", head(available_vars$Variable.Name, 3), "\n")

# Create mapping from base names to full variable names with _count suffix
otu_with_suffix <- paste0(otu_taxa_names, "_count")

tax_mat_nhanes <- available_vars %>%
  filter(Variable.Name %in% otu_with_suffix) %>%
  mutate(
    # Ensure Variable.Description is character and handle NAs
    desc_clean = as.character(ifelse(is.na(Variable.Description) | Variable.Description == "", "", Variable.Description))
  ) %>%
  mutate(
    # Parse taxonomy description
    desc_parts = strsplit(desc_clean, ";"),
    Domain = sapply(desc_parts, function(x) ifelse(length(x) >= 1 && trimws(x[1]) != "", trimws(x[1]), NA_character_)),
    Phylum = sapply(desc_parts, function(x) ifelse(length(x) >= 2 && trimws(x[2]) != "", trimws(x[2]), NA_character_)),
    Class  = sapply(desc_parts, function(x) ifelse(length(x) >= 3 && trimws(x[3]) != "", trimws(x[3]), NA_character_)),
    Order  = sapply(desc_parts, function(x) ifelse(length(x) >= 4 && trimws(x[4]) != "", trimws(x[4]), NA_character_)),
    Family = sapply(desc_parts, function(x) ifelse(length(x) >= 5 && trimws(x[5]) != "", trimws(x[5]), NA_character_)),
    Genus  = sapply(desc_parts, function(x) ifelse(length(x) >= 6 && trimws(x[6]) != "", trimws(x[6]), NA_character_))
  ) %>%
  # Replace "NA" strings with actual NA values
  mutate(
    Domain = ifelse(Domain == "NA" | Domain == "", NA_character_, Domain),
    Phylum = ifelse(Phylum == "NA" | Phylum == "", NA_character_, Phylum),
    Class = ifelse(Class == "NA" | Class == "", NA_character_, Class),
    Order = ifelse(Order == "NA" | Order == "", NA_character_, Order),
    Family = ifelse(Family == "NA" | Family == "", NA_character_, Family),
    Genus = ifelse(Genus == "NA" | Genus == "", NA_character_, Genus)
  ) %>%
  # Convert back to base names for OTU matrix matching
  mutate(otu_base = gsub("_count$", "", Variable.Name)) %>%
  select(otu_base, Domain, Phylum, Class, Order, Family, Genus) %>%
  column_to_rownames("otu_base") %>%
  as.matrix()

cat("✅ Taxonomy table created:", nrow(tax_mat_nhanes), "taxa ×", ncol(tax_mat_nhanes), "taxonomic levels\n")

# Display sample of taxonomy table
cat("\nSample taxonomy entries:\n")
if (nrow(tax_mat_nhanes) > 0) {
  print(head(tax_mat_nhanes, 3))
} else {
  cat("No taxonomy entries found - check variable name matching\n")
}
```

## **3. Create Sample Metadata**

```{r sample_metadata}
cat("Creating sample metadata...\n")

# Read processed demographic data (already has comprehensive variables)
demo_f <- tbl(con, "DEMO_F") %>%
  collect() %>%
  mutate(SEQN = as.character(SEQN)) %>%
  filter(SEQN %in% colnames(otu_mat_counts))

demo_g <- tbl(con, "DEMO_G") %>%
  collect() %>%
  mutate(SEQN = as.character(SEQN)) %>%
  filter(SEQN %in% colnames(otu_mat_counts))

cat("DEMO_F variables:", ncol(demo_f), "| DEMO_G variables:", ncol(demo_g), "\n")

# Combine demographic data using common columns
common_cols <- intersect(names(demo_f), names(demo_g))
cat("Common variables between F and G:", length(common_cols), "\n")

demo_combined <- bind_rows(
  demo_f %>% select(all_of(common_cols)),
  demo_g %>% select(all_of(common_cols))
)

# Create sample metadata dataframe with ALL variables
samples_df_nhanes <- demo_combined %>%
  # Rename SEQN to sample_id for phyloseq
  rename(sample_id = SEQN) %>%
  # Keep ALL variables (no selection/filtering)
  # Set sample_id as rownames for phyloseq
  column_to_rownames("sample_id")

cat("✅ Sample metadata created:", nrow(samples_df_nhanes), "samples ×", ncol(samples_df_nhanes), "variables\n")

# Display all available variables
cat("\nAll sample metadata variables:\n")
all_vars <- colnames(samples_df_nhanes)
cat("Total variables:", length(all_vars), "\n")

# Group variables by type for better display
demographic_vars <- all_vars[grepl("(?i)(gender|age|ethnicity|race|education|income|pir|born|citizen)", all_vars)]
household_vars <- all_vars[grepl("(?i)(household|family|marital)", all_vars)]
survey_vars <- all_vars[grepl("(?i)(weight|wt|stratum|psu|cycle|interview)", all_vars)]
other_vars <- setdiff(all_vars, c(demographic_vars, household_vars, survey_vars))

if (length(demographic_vars) > 0) {
  cat("\nDemographic variables (", length(demographic_vars), "):\n")
  cat("  ", paste(demographic_vars, collapse = ", "), "\n")
}

if (length(household_vars) > 0) {
  cat("\nHousehold variables (", length(household_vars), "):\n")
  cat("  ", paste(household_vars, collapse = ", "), "\n")
}

if (length(survey_vars) > 0) {
  cat("\nSurvey design variables (", length(survey_vars), "):\n")
  cat("  ", paste(survey_vars, collapse = ", "), "\n")
}

if (length(other_vars) > 0) {
  cat("\nOther variables (", length(other_vars), "):\n")
  cat("  ", paste(other_vars, collapse = ", "), "\n")
}

# Verify essential NHANES survey variables are present
essential_survey_vars <- c("WTINT2YR", "WTMEC2YR", "SDMVPSU", "SDMVSTRA")
missing_essential <- setdiff(essential_survey_vars, all_vars)
if (length(missing_essential) == 0) {
  cat("\n✅ All essential NHANES survey design variables are present\n")
} else {
  cat("\n⚠️ Missing essential survey variables:", paste(missing_essential, collapse = ", "), "\n")
}
```

## **4. Create Phylogenetic Tree**

```{r phylogenetic_tree}
cat("Creating taxonomy-based phylogenetic tree...\n")

# Create taxonomy dataframe for tree construction
physeq_tax_df <- tax_mat_nhanes %>%
  as.data.frame() %>%
  rownames_to_column("otu") %>%
  # Clean taxonomic names
  mutate(across(Domain:Genus, ~str_trim(.))) %>%
  mutate(across(Domain:Genus, function(x) {
  x <- gsub("_\\(.*?\\)", "", x)   # Remove annotations like '_(Subgroup_1)'
  x <- gsub("_", " ", x)           # Replace underscores with spaces
    x <- str_trim(x)                 # Trim whitespace
  return(x)
  })) %>%
  # Handle missing values
  mutate(across(c(Domain, Phylum, Class, Order, Family, Genus), ~ {
    x <- gsub("(?i)(Unknown|Unclassified)", "unclassified", .)
    ifelse(is.na(x), "unclassified", x)
  }))

# Create hierarchical path for tree construction
physeq_tax_df$FullPath <- with(
  physeq_tax_df,
  paste(Domain, Phylum, Class, Order, Family, Genus, otu, sep = "/")
)

# Build the data.tree
taxonomy_tree <- data.tree::as.Node(
  physeq_tax_df,
  pathName = "FullPath",
  pathDelimiter = "/"
)

# Sanitize node names
sanitize_node_names <- function(node) {
  node$name <- gsub("[^a-zA-Z0-9._-]+", "_", node$name)
  if (!node$isLeaf) {
    for (child in node$children) {
      sanitize_node_names(child)
    }
  }
}
sanitize_node_names(taxonomy_tree)

# Convert to Newick string
convert_to_newick <- function(node) {
  if (node$isLeaf) {
    return(node$name)
  } else {
    children_newick <- sapply(node$children, convert_to_newick)
    return(paste0("(", paste(children_newick, collapse = ","), ")", node$name))
  }
}
newick_str <- paste0(convert_to_newick(taxonomy_tree), ";")

# Create phylo object
phylo_obj <- ape::read.tree(text = newick_str)
phylo_obj <- ape::multi2di(phylo_obj)

# Fix branch lengths
if (is.null(phylo_obj$edge.length)) {
  phylo_obj$edge.length <- rep(1, nrow(phylo_obj$edge))
}
phylo_obj$edge.length[is.na(phylo_obj$edge.length) | is.infinite(phylo_obj$edge.length)] <- 1

# Prune to match OTU tables
common_tips <- intersect(phylo_obj$tip.label, rownames(otu_mat_counts))
if (length(common_tips) == 0) {
  stop("❌ No overlapping tips found between tree and OTU table")
}
phylo_obj <- ape::keep.tip(phylo_obj, common_tips)

cat("✅ Phylogenetic tree created with", length(phylo_obj$tip.label), "tips\n")
```

## **5. Create phyloseq Objects**

```{r create_phyloseq_objects}
cat("Creating phyloseq objects...\n")

# Create phyloseq components
OTU_counts <- otu_table(otu_mat_counts, taxa_are_rows = TRUE)
OTU_relative <- otu_table(otu_mat_relative, taxa_are_rows = TRUE)
OTU_relative_none <- otu_table(otu_mat_relative_none, taxa_are_rows = TRUE)
OTU_relative_clr <- otu_table(otu_mat_relative_clr, taxa_are_rows = TRUE)
OTU_relative_lognorm <- otu_table(otu_mat_relative_lognorm, taxa_are_rows = TRUE)
TAX <- tax_table(tax_mat_nhanes)
SAMPLES <- sample_data(samples_df_nhanes)

# Create phyloseq objects
cat("Creating ubiome_counts (raw counts, no filtering)...\n")
ubiome_counts <- phyloseq(OTU_counts, TAX, SAMPLES)

cat("Creating ubiome_relative (relative abundance, no filtering)...\n")
ubiome_relative <- phyloseq(OTU_relative, TAX, SAMPLES)

cat("Creating ubiome_relative_none (0.1% prevalence filtered, no transformation)...\n")
ubiome_relative_none <- phyloseq(OTU_relative_none, TAX, SAMPLES)

cat("Creating ubiome_relative_clr (0.1% prevalence filtered, CLR transformed)...\n")
ubiome_relative_clr <- phyloseq(OTU_relative_clr, TAX, SAMPLES)

cat("Creating ubiome_relative_lognorm (0.1% prevalence filtered, log-normal transformed)...\n")
ubiome_relative_lognorm <- phyloseq(OTU_relative_lognorm, TAX, SAMPLES)

# Summary of phyloseq objects
cat("\nSummary of phyloseq objects:\n")
phyloseq_objects <- list(
  "ubiome_counts" = ubiome_counts,
  "ubiome_relative" = ubiome_relative,
  "ubiome_relative_none" = ubiome_relative_none,
  "ubiome_relative_clr" = ubiome_relative_clr,
  "ubiome_relative_lognorm" = ubiome_relative_lognorm
)

for (name in names(phyloseq_objects)) {
  ps <- phyloseq_objects[[name]]
  cat("  ", name, ":\n")
  cat("    Taxa:", ntaxa(ps), "| Samples:", nsamples(ps), "\n")
  cat("    Sample variables:", length(sample_variables(ps)), "\n")
}

cat("\n✅ All 5 phyloseq objects created successfully!\n")
```

## **6. Save phyloseq Objects**

```{r save_phyloseq_objects}
cat("💾 Saving phyloseq objects...\n")

# Save all phyloseq objects
saveRDS(ubiome_counts, file.path(intermediate_files_path, "ubiome_counts.rds"))
saveRDS(ubiome_relative, file.path(intermediate_files_path, "ubiome_relative.rds"))
saveRDS(ubiome_relative_none, file.path(intermediate_files_path, "ubiome_relative_none.rds"))
saveRDS(ubiome_relative_clr, file.path(intermediate_files_path, "ubiome_relative_clr.rds"))
saveRDS(ubiome_relative_lognorm, file.path(intermediate_files_path, "ubiome_relative_lognorm.rds"))

cat("✅ Phyloseq objects saved:\n")
cat("  - ubiome_counts:", file.path(intermediate_files_path, "ubiome_counts.rds"), "\n")
cat("  - ubiome_relative:", file.path(intermediate_files_path, "ubiome_relative.rds"), "\n")
cat("  - ubiome_relative_none:", file.path(intermediate_files_path, "ubiome_relative_none.rds"), "\n")
cat("  - ubiome_relative_clr:", file.path(intermediate_files_path, "ubiome_relative_clr.rds"), "\n")
cat("  - ubiome_relative_lognorm:", file.path(intermediate_files_path, "ubiome_relative_lognorm.rds"), "\n")

# Also save combined list for convenience
phyloseq_objects <- list(
  counts = ubiome_counts,
  relative = ubiome_relative,
  relative_none = ubiome_relative_none,
  relative_clr = ubiome_relative_clr,
  relative_lognorm = ubiome_relative_lognorm
)
saveRDS(phyloseq_objects, file.path(intermediate_files_path, "nhanes_phyloseq_objects_all.rds"))
cat("  - All objects:", file.path(intermediate_files_path, "nhanes_phyloseq_objects_all.rds"), "\n")
```

## **7. Validation and Summary**

```{r validation}
cat("Validation and Summary:\n")

# Function to summarize phyloseq object
summarize_phyloseq <- function(ps, name) {
  cat("\n", name, ":\n")
  cat("  Samples:", nsamples(ps), "\n")
  cat("  Taxa:", ntaxa(ps), "\n")
  cat("  Sample variables:", length(sample_variables(ps)), "\n")
  cat("  Taxonomic ranks:", paste(rank_names(ps), collapse = ", "), "\n")
  
  # Check for negative values (should only be in CLR)
  otu_data <- as.matrix(otu_table(ps))
  has_negative <- any(otu_data < 0, na.rm = TRUE)
  has_zero <- any(otu_data == 0, na.rm = TRUE)
  cat("  Has negative values:", has_negative, "\n")
  cat("  Has zero values:", has_zero, "\n")
  
  # Sample data summary
  cat("  Key sample variables available:\n")
  key_vars <- c("Gender", "AgeGroup", "Ethnicity", "EducationLevel", "cycle")
  for (var in key_vars) {
    if (var %in% sample_variables(ps)) {
      cat("    ✅", var, "\n")
} else {
      cat("    ❌", var, "(missing)\n")
    }
  }
}

# Summarize all objects
summarize_phyloseq(ubiome_counts, "ubiome_counts")
summarize_phyloseq(ubiome_relative, "ubiome_relative") 
summarize_phyloseq(ubiome_relative_none, "ubiome_relative_none")
summarize_phyloseq(ubiome_relative_clr, "ubiome_relative_clr")
summarize_phyloseq(ubiome_relative_lognorm, "ubiome_relative_lognorm")

# Verify sample consistency across objects
sample_consistency <- all(
  identical(sample_names(ubiome_counts), sample_names(ubiome_relative)),
  identical(sample_names(ubiome_counts), sample_names(ubiome_relative_none)),
  identical(sample_names(ubiome_counts), sample_names(ubiome_relative_clr)),
  identical(sample_names(ubiome_counts), sample_names(ubiome_relative_lognorm))
)
cat("\n✅ Sample names consistent across all objects:", sample_consistency, "\n")

# Verify taxa consistency
taxa_consistency <- all(
  identical(taxa_names(ubiome_counts), taxa_names(ubiome_relative)),
  identical(taxa_names(ubiome_counts), taxa_names(ubiome_relative_none)),
  identical(taxa_names(ubiome_counts), taxa_names(ubiome_relative_clr)),
  identical(taxa_names(ubiome_counts), taxa_names(ubiome_relative_lognorm))
)
cat("✅ Taxa names consistent across all objects:", taxa_consistency, "\n")
```
Here is your script converted into clean **Markdown** format with proper section headings, code blocks, and explanatory structure:

---

## **8. Usage Examples**

### 1. **Loading Specific Phyloseq Object**

```{txt}
ubiome_relative_clr <- readRDS(file.path(intermediate_files_path, "ubiome_relative_clr.rds"))
```

### 2. **Loading All Phyloseq Objects**

```{txt}
all_objects <- readRDS(file.path(intermediate_files_path, "nhanes_phyloseq_objects_all.rds"))
ubiome_relative_clr <- all_objects$relative_clr
```

### 3. **Basic Analysis Workflow**

```{txt}
# Alpha diversity (use counts data)
alpha_div <- estimate_richness(ubiome_counts)

# Beta diversity (use transformed data)
ordination <- ordinate(ubiome_relative_clr, method = "PCoA", distance = "euclidean")

# Differential abundance (choose appropriate transformation)
# CLR for compositional models; counts for DESeq2
```

### 4. **Survey-Weighted Analysis Preparation**

```{txt}
# Extract survey design variables
sample_data <- data.frame(sample_data(ubiome_relative_clr))
weights <- sample_data$Weight_MEC_Exam
strata <- sample_data$Masked_Variance_Pseudo_Stratum
clusters <- sample_data$Masked_Variance_Pseudo_PSU
```

---

## ✅ **9. Completion Summary**

### **Created Objects**

1. `ubiome_counts` – Original count data (for alpha diversity)
2. `ubiome_relative` – Relative abundance (for beta diversity)
3. `ubiome_relative_none` – 0.1% prevalence filtered, no transformation
4. `ubiome_relative_clr` – CLR-transformed (0.1% filtered)
5. `ubiome_relative_lognorm` – Log-normal transformed (0.1% filtered)

---

### **Output Files**

* **Individual RDS files:** Located in `intermediate_files_path`
* **Combined object file:** `nhanes_phyloseq_objects_all.rds`
* **Taxonomy table:** `tax_mat_nhanes.csv`
* **Sample metadata:** `samples_df_nhanes.csv`

---

### **Integration with Analysis Pipeline**

* Use **`ubiome_counts`** for: Alpha diversity, DESeq2
* Use **`ubiome_relative`** for: Beta diversity, compositional summaries
* Use **`ubiome_relative_none`** for: Raw baseline relative abundance
* Use **`ubiome_relative_clr`** for: CLR-based and distance-based analysis
* Use **`ubiome_relative_lognorm`** for: statistical tests
* **All objects** include: survey weights and design variables

---

### **Next Steps**

1. Proceed to **Chapter 3**: GOLD database integration
2. Perform downstream statistical analyses using the correct transformation
3. Apply survey weights for population-representative inference


## **9. Completion Summary**

```{r Disconnect}
# Disconnect from database
dbDisconnect(con)
cat("\nDatabase connection closed.\n")
```

## **Data Normalization Guidelines**

### **Should I normalize data before alpha-diversity analysis?**

**No.** Use **original count data (ubiome_counts) for alpha-diversity analyses.** Rarefying data for normalization, as often suggested in pipelines like QIIME, is statistically flawed.

### **Recommended Reading:**
- Gotelli & Colwell (2001)
- ["Waste Not, Want Not…"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4370911/)

### **Transformation Guide:**
- **ubiome_counts**: Alpha diversity, DESeq2, original statistical tests
- **ubiome_relative**: Beta diversity, compositional data analysis (recommended)
- **ubiome_relative_none**: Raw relative abundance baseline
- **ubiome_relative_clr**: Beta diversity, compositional data analysis (recommended)
- **ubiome_relative_lognorm**: Standard statistical tests, linear modeling

(Source: [Phyloseq FAQ](https://www.bioconductor.org/packages/release/bioc/vignettes/phyloseq/inst/doc/phyloseq-FAQ.html))