---
title: "Verify NHANES Oral Microbiome Transformation Pipeline Completeness"
subtitle: "Check 4 Transformations (none, hellinger, clr, lognorm) and Metadata Coverage"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
date: "`r Sys.Date()`"
---

```{r setup, message=FALSE, warning=FALSE}
# Load packages
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(purrr)
library(knitr)
library(stringr)

# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Overview

This document verifies the completeness of the NHANES oral microbiome transformation pipeline after implementing the corrected transformations:

- **4 Transformations**: `none`, `hellinger`, `clr`, `lognorm`
- **Universal Naming**: All output tables use `DADA2RSV_<taxalevel>_RELATIVE_<cycle>_<method>` format
- **Input Sources**: RELATIVE tables for none/hellinger, COUNT tables for clr/lognorm
- **Pseudocount**: ε = 0.001 applied only to COUNT matrices

## 1. Connect to Database

```{r connect}
# Use relative path from script directory
db_path <- "../../data/00_nhanes_omp_transformed_db/nhanes_oral_transformed_complete.sqlite"

# Check if database exists
if (!file.exists(db_path)) {
  stop("Database not found at: ", db_path)
}

cat("✓ Database found:", db_path, "\n")
cat("  Size:", round(file.size(db_path) / 1024^3, 2), "GB\n")

con <- dbConnect(SQLite(), db_path)
```

## 2. Verify Transformation Tables

Expected: 40 transformed tables (10 taxonomic levels × 4 transformations)

```{r check-transformations}
# Get all table names
all_tables <- dbListTables(con)

# Define expected transformation patterns
taxonomic_levels <- c("GENUS", "FAMILY", "ORDER", "CLASS", "PHYLUM")
cycles <- c("F", "G")
transformations <- c("none", "hellinger", "clr", "lognorm")

# Generate expected table names (all use RELATIVE format)
expected_tables <- expand.grid(
  level = taxonomic_levels,
  cycle = cycles,
  transform = transformations,
  stringsAsFactors = FALSE
) %>%
  mutate(
    table_name = paste0("DADA2RSV_", level, "_RELATIVE_", cycle, "_", transform)
  ) %>%
  pull(table_name)

cat("Expected transformed tables:", length(expected_tables), "\n")

# Check which tables exist
existing_transformed <- all_tables[str_detect(all_tables, "DADA2RSV_.*_RELATIVE_.*_(none|hellinger|clr|lognorm)$")]
missing_tables <- setdiff(expected_tables, existing_transformed)
extra_tables <- setdiff(existing_transformed, expected_tables)

cat("Found transformed tables:", length(existing_transformed), "\n")
cat("Missing tables:", length(missing_tables), "\n")
cat("Extra tables:", length(extra_tables), "\n")

# Create summary table
transformation_summary <- data.frame(
  Transformation = transformations,
  Expected = length(taxonomic_levels) * length(cycles),
  Found = sapply(transformations, function(t) {
    sum(str_detect(existing_transformed, paste0("_", t, "$")))
  }),
  stringsAsFactors = FALSE
) %>%
  mutate(
    Complete = Found == Expected,
    Status = ifelse(Complete, "✓ Complete", "✗ Missing")
  )

kable(transformation_summary, caption = "Transformation Completeness Summary")

if (length(missing_tables) > 0) {
  cat("\n⚠️ Missing tables:\n")
  cat(paste("  -", missing_tables), sep = "\n")
}

if (length(extra_tables) > 0) {
  cat("\n⚠️ Extra tables found:\n")
  cat(paste("  -", extra_tables), sep = "\n")
}
```

## 3. Sample Data from Each Transformation Type

```{r sample-data}
# Sample one table from each transformation type
sample_tables <- c(
  "DADA2RSV_GENUS_RELATIVE_F_none",
  "DADA2RSV_GENUS_RELATIVE_F_hellinger", 
  "DADA2RSV_GENUS_RELATIVE_F_clr",
  "DADA2RSV_GENUS_RELATIVE_F_lognorm"
)

sample_results <- list()

for (tbl in sample_tables) {
  if (tbl %in% all_tables) {
    df <- dbReadTable(con, tbl)
    transform_type <- str_extract(tbl, "(none|hellinger|clr|lognorm)$")
    
    # Get basic stats
    n_samples <- nrow(df)
    n_taxa <- ncol(df) - 1  # Exclude SEQN column
    
    # Sample some values (excluding SEQN column)
    sample_values <- as.numeric(df[1, -1][1:5])  # First 5 taxa values
    
    sample_results[[transform_type]] <- list(
      table = tbl,
      samples = n_samples,
      taxa = n_taxa,
      min_val = min(sample_values, na.rm = TRUE),
      max_val = max(sample_values, na.rm = TRUE),
      range_check = case_when(
        transform_type %in% c("none", "hellinger") ~ "Expected [0,1]",
        transform_type %in% c("clr", "lognorm") ~ "Expected (-∞,+∞)"
      )
    )
  }
}

# Create sample data summary
sample_df <- map_dfr(sample_results, ~data.frame(
  Samples = .x$samples,
  Taxa = .x$taxa,
  Min_Value = round(.x$min_val, 4),
  Max_Value = round(.x$max_val, 4),
  Expected_Range = .x$range_check
), .id = "Transformation")

kable(sample_df, caption = "Sample Data Characteristics by Transformation")
```

## 4. Check Derived Variables and Metadata

```{r vars-check}
# Updated variable list based on current pipeline
vars_to_check <- c(
  # Laboratory biomarkers
  "LBXP1", "LBXP2", "LBDP3", "LBDRFO", "LBXSF6SI", "LBXSCK",
  "LBXFOLSI", "LBXPS4", "LBXTBM", "LBDFOT", "LBXMMASI", "URX1DC",
  
  # Anthropometric measures  
  "BMXSUB", "BMXTRI",
  
  # Health outcomes
  "ASTHMA", "BRONCHITIS", "EMPHYSEMA", "ANGINA", "HEART_FAILURE", 
  "HEART_ATTACK", "STROKE", "CHD", "CVD",
  
  # Cancer history
  "CANCER_BREAST", "CANCER_COLON", "CANCER_LUNG", "CANCER_ESOPHAGEAL",
  "CANCER_PROSTATE", "CANCER_MOUTH",
  
  # Other health conditions
  "DIABETES",
  
  # Derived demographic variables
  "AGE_SQUARED", "RIAGENDR_01", "BORN_INUSA",
  
  # Education indicators
  "EDUCATION_LESS9", "EDUCATION_9_11", "EDUCATION_AA", "EDUCATION_COLLEGEGRAD",
  
  # Ethnicity indicators
  "ETHNICITY_MEXICAN", "ETHNICITY_OTHERHISPANIC", "ETHNICITY_OTHER", 
  "ETHNICITY_NONHISPANICBLACK"
)

# Find which tables contain each variable
fields_by_tbl <- setNames(
  lapply(all_tables, function(tbl) dbListFields(con, tbl)),
  all_tables
)

# Create mapping
var_mapping <- map_dfr(vars_to_check, function(v) {
  present <- names(fields_by_tbl)[
    vapply(fields_by_tbl, function(flds) v %in% flds, logical(1))
  ]
  
  tibble(
    Variable = v,
    Found_In_Tables = ifelse(length(present) == 0, "NOT FOUND", 
                            paste(present, collapse = ", ")),
    Table_Count = length(present)
  )
})

kable(var_mapping, caption = "Variable Location Mapping")

# Summary of variable availability
var_summary <- var_mapping %>%
  summarise(
    Total_Variables = n(),
    Found_Variables = sum(Table_Count > 0),
    Missing_Variables = sum(Table_Count == 0),
    Completion_Rate = round(Found_Variables / Total_Variables * 100, 1)
  )

cat("\nVariable Availability Summary:\n")
cat("  Total variables checked:", var_summary$Total_Variables, "\n")
cat("  Found in database:", var_summary$Found_Variables, "\n") 
cat("  Missing from database:", var_summary$Missing_Variables, "\n")
cat("  Completion rate:", var_summary$Completion_Rate, "%\n")
```

## 5. Check Metadata Coverage

```{r metadata-check}
# Read metadata table
meta_raw <- dbReadTable(con, "variable_names_epcf")

cat("Metadata table summary:\n")
cat("  Total metadata rows:", nrow(meta_raw), "\n")
cat("  Unique variables:", length(unique(meta_raw$Variable.Name)), "\n")
cat("  Unique tables:", length(unique(meta_raw$Data.File.Name)), "\n")

# Check which variables have metadata
meta_coverage <- var_mapping %>%
  mutate(
    Has_Metadata = Variable %in% meta_raw$Variable.Name
  ) %>%
  filter(Table_Count > 0)  # Only check variables that exist in tables

missing_metadata <- meta_coverage %>%
  filter(!Has_Metadata) %>%
  select(Variable, Found_In_Tables)

if (nrow(missing_metadata) > 0) {
  cat("\n⚠️ Variables missing from metadata:\n")
  kable(missing_metadata, caption = "Variables in Tables but Missing from Metadata")
} else {
  cat("\n✅ All variables found in tables are documented in metadata!\n")
}

# Check transformation table metadata coverage
transformation_meta <- meta_raw %>%
  filter(str_detect(Data.File.Name, "DADA2RSV_.*_RELATIVE_.*(none|hellinger|clr|lognorm)$")) %>%
  mutate(
    Transformation = str_extract(Data.File.Name, "(none|hellinger|clr|lognorm)$")
  ) %>%
  count(Transformation, name = "Metadata_Entries")

cat("\n📝 Transformation metadata coverage:\n")
kable(transformation_meta, caption = "Metadata Entries by Transformation Type")
```

## 6. Database Schema Overview

```{r schema-overview}
# Count different table types
table_types <- tibble(
  Table_Name = all_tables
) %>%
  mutate(
    Type = case_when(
      str_detect(Table_Name, "DADA2RSV_.*_RELATIVE_.*(none|hellinger|clr|lognorm)$") ~ "Transformed Microbiome",
      str_detect(Table_Name, "DADA2RSV_.*_(RELATIVE|COUNT)_[FG]$") ~ "Original Microbiome", 
      str_detect(Table_Name, "DEMO_[FG]$") ~ "Demographics",
      str_detect(Table_Name, "^[A-Z]+_[FG]$") ~ "NHANES Data",
      Table_Name == "variable_names_epcf" ~ "Metadata",
      TRUE ~ "Other"
    )
  ) %>%
  count(Type, name = "Count")

kable(table_types, caption = "Database Schema Overview")

cat("\nTotal tables in database:", length(all_tables), "\n")
```

## 7. Quality Checks

```{r quality-checks}
# Check for SEQN consistency across a few tables
seqn_check <- list()

check_tables <- c(
  "DADA2RSV_GENUS_RELATIVE_F_none",
  "DADA2RSV_GENUS_RELATIVE_F_clr", 
  "DEMO_F"
)

for (tbl in check_tables) {
  if (tbl %in% all_tables) {
    seqn_ids <- dbGetQuery(con, paste("SELECT SEQN FROM", tbl))$SEQN
    seqn_check[[tbl]] <- list(
      count = length(seqn_ids),
      unique_count = length(unique(seqn_ids)),
      has_duplicates = length(seqn_ids) != length(unique(seqn_ids))
    )
  }
}

seqn_df <- map_dfr(seqn_check, ~data.frame(
  Total_Rows = .x$count,
  Unique_SEQN = .x$unique_count,
  Has_Duplicates = .x$has_duplicates
), .id = "Table")

kable(seqn_df, caption = "SEQN Consistency Check")

# Check transformation naming convention compliance
naming_check <- existing_transformed %>%
  tibble(Table = .) %>%
  mutate(
    Follows_Convention = str_detect(Table, "^DADA2RSV_(GENUS|FAMILY|ORDER|CLASS|PHYLUM)_RELATIVE_[FG]_(none|hellinger|clr|lognorm)$"),
    Issue = ifelse(!Follows_Convention, "❌ Non-standard naming", "✅ Correct naming")
  )

naming_issues <- naming_check %>% filter(!Follows_Convention)

if (nrow(naming_issues) > 0) {
  cat("\n⚠️ Naming convention issues found:\n")
  kable(naming_issues, caption = "Tables with Naming Issues")
} else {
  cat("\n✅ All transformation tables follow the correct naming convention!\n")
}
```

## 8. Summary and Recommendations

```{r summary}
cat("PIPELINE VERIFICATION SUMMARY\n")
cat("================================\n\n")

cat("✅ SUCCESSES:\n")
cat("  • Database accessible and properly sized\n")
cat("  • 4 transformation methods implemented:", paste(transformations, collapse = ", "), "\n")
cat("  • Universal RELATIVE naming convention applied\n")
cat("  • Expected number of transformed tables:", length(existing_transformed), "out of", length(expected_tables), "\n")

if (length(missing_tables) == 0 && nrow(missing_metadata) == 0) {
  cat("  • All transformations complete and documented\n")
}

if (length(missing_tables) > 0 || nrow(missing_metadata) > 0) {
  cat("\n⚠️ ISSUES TO ADDRESS:\n")
  
  if (length(missing_tables) > 0) {
    cat("  • Missing", length(missing_tables), "transformation tables\n")
  }
  
  if (nrow(missing_metadata) > 0) {
    cat("  • Missing metadata for", nrow(missing_metadata), "variables\n")
  }
}

cat("\nNEXT STEPS:\n")
cat("  1. If any tables missing, re-run transformation script\n") 
cat("  2. If metadata missing, run nhanes_db_filling_missing_data.R\n")
cat("  3. Proceed to schema structure generation (24 files expected)\n")
cat("  4. Continue to association analysis pipeline\n")

completion_rate <- round(length(existing_transformed) / length(expected_tables) * 100, 1)
cat("\nOverall completion rate:", completion_rate, "%\n")
```

```{r cleanup}
dbDisconnect(con)
cat("\n✅ Database connection closed.\n")
```

---

**Generated on:** `r Sys.time()`  
**Pipeline Version:** 4-Transformation Implementation (none, hellinger, clr, lognorm)  
**Database:** `nhanes_oral_transformed_complete.sqlite`
